---
title: "Data preparation"
subtitle: "Market justice preferences with ISSP (1987-2019)"
author: "Tomás Urzúa"
date: today
lang: es
fontsize: 12pt
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    toc-expand: 2
    toc-title: Contents
    number-sections: true
    number-depth: 3
    theme:
      - cosmo
      - edumer_html.scss
    code-link: true
    code-fold: true
    title-block-banner: true
  pdf:
    number-sections: true
    number-depth: 3
editor_options: 
  chunk_output_type: console
---

# Presentation

Preparation for 1987, 1992, 1999, 2009 and 2019 database from [Social Inequality Module of International Social Survey Programme (ISSP)](https://www.gesis.org/en/issp/data-and-documentation/social-inequality).

```{r}
#| label: set
#| echo: false
#| message: false
#| warning: false

library(knitr)
knitr::opts_chunk$set(echo = TRUE, include = TRUE, warning = FALSE, message = FALSE)

table_format <- if(is_html_output()) {
  "html"
} else if(is_latex_output()) {
  "latex"
}
table_format2 <- if(is_html_output()) {
  T
} else if(is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")
```


## Libraries

```{r}
#| label: packages
#| include: true

if (! require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse,
               car,
               sjmisc,
               here,
               sjlabelled,
               SciViews,
               naniar,
               haven,
               countrycode,
               ggrepel)


library(occupar)
options(scipen=999)
rm(list = ls())
```

## Data

```{r}
#| label: data
#| echo: true
#| output: false

issp87 <- read_dta(here("input/data/original/issp_1987.dta"))
issp92 <- read_dta(here("input/data/original/issp_1992.dta"))
issp99 <- read_dta(here("input/data/original/issp_1999.dta"))
issp09 <- read_dta(here("input/data/original/issp_2009.dta"))
issp19 <- read_dta(here("input/data/original/issp_2019.dta"))
issp_cumu <- read_dta(here("input/data/original/issp_cumulative.dta"))
```

# ISSP 1987

## Select

```{r}

db_proc87 <- issp87 %>%
  select(id = v2,
         country = v3,
         sex = v83,
         age = v82,
         income = v94,
         ess = v65,
         nmerit_wealthy_fam = v4,
         nmerit_educated_parents = v5,
         merit_ambition = v7,
         merit_effort = v9,
         nmerit_contacts = v10,
         nmerit_political_connection = v11,
         nmerit_race = v12,
         nmerit_gender = v15,
         isco68 = v75,
         selfemp = v77,
         wrksup = v78,
         weight = v107) %>%
  janitor::clean_names()

```


##  Recode and transform

### Year

```{r}
db_proc87$year <- 1987
db_proc87$year <- as.numeric(db_proc87$year)
```

### Country

```{r}
frq(db_proc87$country)
db_proc87 <- db_proc87 %>%
  mutate(country = case_when(
    country == 1  ~ "Australia",
    country == 2  ~ "Germany",
    country == 3  ~ "United Kingdom",
    country == 4  ~ "United States",
    country == 5  ~ "Austria",
    country == 6  ~ "Hungary",
    country == 7  ~ "Netherlands",
    country == 8  ~ "Italy",
    country == 11 ~ "Switzerland",
    country == 12 ~ "Poland",
    TRUE ~ NA_character_
  ))
db_proc87$iso_country <- countrycode::countrycode(db_proc87$country, origin = "country.name", destination = "iso3c")
frq(db_proc87$country); frq(db_proc87$iso_country)

```

### Sex
```{r}
frq(db_proc87$sex)
db_proc87$sex <- set_na(db_proc87$sex, na = c(9), drop.levels = F)
db_proc87 <- db_proc87 %>%
mutate(sex = if_else(sex == 1, 'Male', 'Female'),
       sex = factor(sex, levels = c("Male", "Female")))
frq(db_proc87$sex)
```

### Age
```{r}
summary(db_proc87$age)
db_proc87$age <- set_na(db_proc87$age, na = c(99), drop.levels = F)
db_proc87$age <- as.numeric(db_proc87$age)
```

### Education

```{r}
degree87 <- issp_cumu %>%
  filter(year == 1987) %>%
  select(id = caseid, degree = DEGREE)

db_proc87 <- db_proc87 %>%
  left_join(degree87, by = "id")

frq(db_proc87$degree)
db_proc87$degree <- set_na(db_proc87$degree, na = c(-9), drop.levels = F)
db_proc87 <- db_proc87 %>%
  mutate(educ = case_when(degree == 0 ~ "Incomplete primary or less",
                            degree == 1 ~ "Complete primary",
                            degree == 2 ~ "Incomplete secondary",
                            degree == 3 ~ "Complete secondary",
                            degree == 4 ~ "Incomplete university",
                            degree == 5 ~ "Complete university",
                            TRUE ~ NA_character_),
           educ = factor(educ,
                           levels = c("Incomplete primary or less",
                           "Complete primary",
                           "Incomplete secondary",
                           "Complete secondary",
                           "Incomplete university",
                           "Complete university")))

db_proc87 <- db_proc87 %>%
  mutate(educ_dic = case_when(degree %in% c(0:4) ~ 'Less than universitary',
                              degree == 5 ~ 'Universitary',
                              TRUE ~ NA_character_),
         educ_dic = factor(educ_dic, levels = c("Universitary", "Less than universitary")))
frq(db_proc87$educ)
frq(db_proc87$educ_dic)
```

### Income
```{r}

# ntile function without NAs
ntile_zero_na <- function(x, n = 10) {
  out   <- rep(NA_real_, length(x))
  is_na <- is.na(x)
  is0   <- !is_na & x == 0
  pos   <- !is_na & x > 0
  out[is0] <- 0
  if (any(pos)) out[pos] <- ntile(x[pos], n)
  out
}

summary(as.numeric(db_proc87$income))
db_proc87 <- db_proc87 %>%
  mutate(
    income = as.numeric(income),
    income = if_else(country == "Poland" & income == 0, NA_real_, income), # 0 = self-employed (not zero income) in Poland
    income = car::recode(income, "999990:999999 = NA"),
    income_decil = ntile_zero_na(income, 10)
  )

db_proc87 %>%
  group_by(country) %>%
  summarise(n = n(), n_valid = sum(!is.na(income_decil)), .groups = "drop")
frq(db_proc87$income_decil)

```

### Subjective social status
```{r}
frq(db_proc87$ess)
db_proc87$ess <- set_na(db_proc87$ess, na = c(97,98,99), drop.levels = F)
db_proc87$ess <- as.numeric(db_proc87$ess)
frq(db_proc87$ess)
```

### Meritocratic variables
```{r}
frq(db_proc87$merit_effort)
frq(db_proc87$merit_ambition)

db_proc87 <- db_proc87 %>%
  mutate(
    across(
      .cols = c(merit_effort, merit_ambition),
      .fns = ~ set_na(., na = c(0,7,8,9), drop.levels = F)
    )
  ) # ok

db_proc87$merit_effort <- sjmisc::rec(db_proc87$merit_effort,rec = "rev")
db_proc87$merit_ambition <- sjmisc::rec(db_proc87$merit_ambition,rec = "rev")

frq(db_proc87$merit_effort)
frq(db_proc87$merit_ambition)
```

### Non-meritocratic variables
```{r}
frq(db_proc87$nmerit_wealthy_fam)
frq(db_proc87$nmerit_contacts)
frq(db_proc87$nmerit_educated_parents)
frq(db_proc87$nmerit_political_connection)
frq(db_proc87$nmerit_race)
frq(db_proc87$nmerit_gender)

db_proc87 <- db_proc87 %>%
  mutate(
    across(
      .cols = c(nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents,
                nmerit_political_connection, nmerit_race, nmerit_gender),
      .fns = ~ set_na(., na = c(0,7,8,9), drop.levels = F)
    )
  ) # ok

db_proc87$nmerit_wealthy_fam <- sjmisc::rec(db_proc87$nmerit_wealthy_fam,rec = "rev")
db_proc87$nmerit_contacts <- sjmisc::rec(db_proc87$nmerit_contacts,rec = "rev")
db_proc87$nmerit_educated_parents <- sjmisc::rec(db_proc87$nmerit_educated_parents,rec = "rev")
db_proc87$nmerit_political_connection <- sjmisc::rec(db_proc87$nmerit_political_connection,rec = "rev")
db_proc87$nmerit_race <- sjmisc::rec(db_proc87$nmerit_race,rec = "rev")
db_proc87$nmerit_gender <- sjmisc::rec(db_proc87$nmerit_gender,rec = "rev")

frq(db_proc87$nmerit_wealthy_fam)
frq(db_proc87$nmerit_contacts)
frq(db_proc87$nmerit_educated_parents)
frq(db_proc87$nmerit_political_connection)
frq(db_proc87$nmerit_race)
frq(db_proc87$nmerit_gender)

```

### Market justice healthcare
```{r}
# No disponible en ISSP 1987. El ítem "the rich can buy better health care
# than the poor" fue introducido en ISSP 1999 (Social Inequality III).
db_proc87$just_health <- NA_real_
```

### Market justice in education
```{r}
# No disponible en ISSP 1987, misma razón que just_health.
db_proc87$just_educ <- NA_real_
```

### Perceived inequality
```{r}
# No disponible en ISSP 1987. El ítem del diagrama de pirámide social se
# incorporó recién en ISSP 1992.
db_proc87$perc_inequality <- NA_real_
```

### Class

```{r}
db_proc87 <- db_proc87 %>%
  mutate(
    # El codebook solo confirma ISCO68 de 3 dígitos para Alemania, Austria y
    # Suiza (Nota 1); el resto de países usan escalas propias no ISCO.
    isco68 = case_when(country %in% c("Germany", "Austria", "Switzerland") ~ as.numeric(isco68),
                        TRUE ~ NA_real_),
    isco68 = set_na(isco68, na = c(0,999997,999998,999999), drop.levels = F),
    isco88_4 = isco68to88(isco68),
    selfemp = set_na(selfemp, na = c(0,9), drop.levels = F),
    selfemp_bin = if_else(selfemp %in% c(1,2,3), 1, 0), # incl. "self-employed with employees" y (Italia) "self-employed at someone's account"
    nemploy_rec = NA_real_, # no existe variable de número de empleados en 1987; isco88toEGP() asume 0
    egp5 = isco88toEGP(isco88_4, n.employees = nemploy_rec,
                        self.employed = selfemp_bin, n.classes = 5)
  )

frq(db_proc87$egp5)
```

## Save

```{r}
db_proc87 <- db_proc87 %>%
  select(year, country, iso_country, sex, age, educ, educ_dic, income_decil, ess, just_health, just_educ, perc_inequality, egp5, merit_effort, merit_ambition, nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents, nmerit_political_connection, nmerit_race, nmerit_gender, weight)

save(db_proc87, file = here("input/data/proc/db_proc87.RData"))
```

# ISSP 1992

## Select

```{r}

db_proc92 <- issp92 %>%
  select(id = v2,
         country = v3,
         sex = v99,
         age = v100,
         aus117:rp117,
         ess = v73,
         nmerit_wealthy_fam = v4,
         nmerit_educated_parents = v5,
         merit_ambition = v7,
         merit_effort = v9,
         nmerit_contacts = v10,
         nmerit_political_connection = v11,
         nmerit_race = v12,
         nmerit_gender = v15,
         perc_inequality = v90,
         isco68 = v106,
         aus106, pl106, slo106, rus106, nz106,
         selfemp = v110,
         nemploy = v111,
         aus111, d111, i111, n111, rp111,
         wrksup = v107,
         nwrksup = v108,
         weight = v176) %>%
  janitor::clean_names()

```

##  Recode and transform

### Year

```{r}
db_proc92$year <- 1992
db_proc92$year <- as.numeric(db_proc92$year)
```

### Country

```{r}
frq(db_proc92$country)
db_proc92 <- db_proc92 %>%
  mutate(country = case_when(
    country == 1  ~ "Australia",
    country == 2  ~ "Germany",
    country == 3  ~ "Germany",
    country == 4  ~ "United Kingdom",
    country == 5  ~ "United States",
    country == 6  ~ "Austria",
    country == 7  ~ "Hungary",
    country == 8  ~ "Italy",
    country == 9  ~ "Norway",
    country == 10 ~ "Sweden",
    country == 11 ~ "Czechia",
    country == 12 ~ "Slovenia",
    country == 13 ~ "Poland",
    country == 14 ~ "Bulgaria",
    country == 15 ~ "Russia",
    country == 16 ~ "New Zealand",
    country == 17 ~ "Canada",
    country == 18 ~ "Philippines",
    TRUE ~ NA_character_
  ))
db_proc92$iso_country <- countrycode::countrycode(db_proc92$country, origin = "country.name", destination = "iso3c")
frq(db_proc92$country); frq(db_proc92$iso_country)

```

### Sex
```{r}
frq(db_proc92$sex)
db_proc92$sex <- set_na(db_proc92$sex, na = c(9), drop.levels = F)
db_proc92 <- db_proc92 %>%
mutate(sex = if_else(sex == 1, 'Male', 'Female'),
       sex = factor(sex, levels = c("Male", "Female")))
frq(db_proc92$sex)
```

### Age
```{r}
summary(db_proc92$age)
db_proc92$age <- set_na(db_proc92$age, na = c(99), drop.levels = F)
db_proc92$age <- as.numeric(db_proc92$age)
```

### Education

```{r}
degree92 <- issp_cumu %>%
  filter(year == 1992) %>%
  select(id = caseid, degree = DEGREE)

db_proc92 <- db_proc92 %>%
  left_join(degree92, by = "id")

frq(db_proc92$degree)
db_proc92$degree <- set_na(db_proc92$degree, na = c(-9), drop.levels = F)
db_proc92 <- db_proc92 %>%
  mutate(educ = case_when(degree == 0 ~ "Incomplete primary or less",
                            degree == 1 ~ "Complete primary",
                            degree == 2 ~ "Incomplete secondary",
                            degree == 3 ~ "Complete secondary",
                            degree == 4 ~ "Incomplete university",
                            degree == 5 ~ "Complete university",
                            TRUE ~ NA_character_),
           educ = factor(educ,
                           levels = c("Incomplete primary or less",
                           "Complete primary",
                           "Incomplete secondary",
                           "Complete secondary",
                           "Incomplete university",
                           "Complete university")))

db_proc92 <- db_proc92 %>%
  mutate(educ_dic = case_when(degree %in% c(0:4) ~ 'Less than universitary',
                              degree == 5 ~ 'Universitary',
                              TRUE ~ NA_character_),
         educ_dic = factor(educ_dic, levels = c("Universitary", "Less than universitary")))
frq(db_proc92$educ)
frq(db_proc92$educ_dic)
```



### Income
```{r}

# ntile function without NAs
ntile_zero_na <- function(x, n = 10) {
  out   <- rep(NA_real_, length(x))
  is_na <- is.na(x)
  is0   <- !is_na & x == 0
  pos   <- !is_na & x > 0
  out[is0] <- 0
  if (any(pos)) out[pos] <- ntile(x[pos], n)
  out
}

db_proc92 <- db_proc92 %>%
  mutate_at(vars(aus117:rp117), ~ as.numeric(.)) %>%
  mutate_at(vars(aus117:rp117), funs(car::recode(., "999997 = NA; 999998 = NA; 999999 = NA"))) %>%
  mutate(
    income_personal = do.call(coalesce, across(aus117:rp117)),
    income_decil    = ntile_zero_na(income_personal, 10)
  )

db_proc92 %>%
  mutate(n_non_na = rowSums(!is.na(pick(aus117:rp117)))) %>%
  count(n_non_na)
frq(db_proc92$income_decil)

```

### Subjective social status
```{r}
frq(db_proc92$ess)
db_proc92$ess <- set_na(db_proc92$ess, na = c(97,98,99), drop.levels = F)
db_proc92$ess <- as.numeric(db_proc92$ess)
frq(db_proc92$ess)
```

### Meritocratic variables
```{r}
frq(db_proc92$merit_effort)
frq(db_proc92$merit_ambition)

db_proc92 <- db_proc92 %>%
  mutate(
    across(
      .cols = c(merit_effort, merit_ambition),
      .fns = ~ set_na(., na = c(8,9), drop.levels = F)
    )
  ) # ok

db_proc92$merit_effort <- sjmisc::rec(db_proc92$merit_effort,rec = "rev")
db_proc92$merit_ambition <- sjmisc::rec(db_proc92$merit_ambition,rec = "rev")

frq(db_proc92$merit_effort)
frq(db_proc92$merit_ambition)
```

### Non-meritocratic variables
```{r}
frq(db_proc92$nmerit_wealthy_fam)
frq(db_proc92$nmerit_contacts)
frq(db_proc92$nmerit_educated_parents)
frq(db_proc92$nmerit_political_connection)
frq(db_proc92$nmerit_race)
frq(db_proc92$nmerit_gender)

db_proc92 <- db_proc92 %>%
  mutate(
    across(
      .cols = c(nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents,
                nmerit_political_connection, nmerit_race, nmerit_gender),
      .fns = ~ set_na(., na = c(8,9), drop.levels = F)
    )
  ) # ok

db_proc92$nmerit_wealthy_fam <- sjmisc::rec(db_proc92$nmerit_wealthy_fam,rec = "rev")
db_proc92$nmerit_contacts <- sjmisc::rec(db_proc92$nmerit_contacts,rec = "rev")
db_proc92$nmerit_educated_parents <- sjmisc::rec(db_proc92$nmerit_educated_parents,rec = "rev")
db_proc92$nmerit_political_connection <- sjmisc::rec(db_proc92$nmerit_political_connection,rec = "rev")
db_proc92$nmerit_race <- sjmisc::rec(db_proc92$nmerit_race,rec = "rev")
db_proc92$nmerit_gender <- sjmisc::rec(db_proc92$nmerit_gender,rec = "rev")

frq(db_proc92$nmerit_wealthy_fam)
frq(db_proc92$nmerit_contacts)
frq(db_proc92$nmerit_educated_parents)
frq(db_proc92$nmerit_political_connection)
frq(db_proc92$nmerit_race)
frq(db_proc92$nmerit_gender)

```

### Market justice healthcare
```{r}
# No disponible en ISSP 1992 (Social Inequality II). El ítem "the rich can buy
# better health care than the poor" fue introducido en ISSP 1999 (Social
# Inequality III).
db_proc92$just_health <- NA_real_
```

### Market justice in education
```{r}
# No disponible en ISSP 1992, misma razón que just_health.
db_proc92$just_educ <- NA_real_
```

### Perceived inequality
```{r}
frq(db_proc92$perc_inequality)
db_proc92$perc_inequality <- set_na(db_proc92$perc_inequality, na = c(0,8,9), drop.levels = F)
db_proc92$perc_inequality <- sjmisc::rec(db_proc92$perc_inequality,rec = "rev")
frq(db_proc92$perc_inequality)
```

### Class

```{r}
db_proc92 <- db_proc92 %>%
  mutate(
    isco68 = set_na(isco68, na = c(0,9997,9998,9999), drop.levels = F),
    isco88_conv = isco68to88(isco68),
    isco88_cty = coalesce(
      na_if(as.numeric(aus106), 0),
      na_if(as.numeric(pl106), 0),
      na_if(as.numeric(slo106), 0),
      na_if(as.numeric(rus106), 0),
      na_if(as.numeric(nz106), 0)
    ),
    isco88_4 = coalesce(isco88_conv, isco88_cty),
    selfemp = set_na(selfemp, na = c(0,9), drop.levels = F),
    selfemp_bin = if_else(selfemp == 1, 1, 0),
    nemploy = set_na(nemploy, na = c(99), drop.levels = F),
    nemploy_cty = coalesce(
      na_if(as.numeric(aus111), 0),
      na_if(as.numeric(d111), 0),
      na_if(as.numeric(i111), 0),
      na_if(as.numeric(n111), 0),
      na_if(as.numeric(rp111), 0)
    ),
    nemploy_rec = coalesce(as.numeric(nemploy), nemploy_cty),
    egp5 = isco88toEGP(isco88_4, n.employees = nemploy_rec,
                        self.employed = selfemp_bin, n.classes = 5)
  )

frq(db_proc92$egp5)
```

## Save

```{r}
db_proc92 <- db_proc92 %>%
  select(year, country, iso_country, sex, age, educ, educ_dic, income_decil, ess, just_health, just_educ, perc_inequality, egp5, merit_effort, merit_ambition, nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents, nmerit_political_connection, nmerit_race, nmerit_gender, weight)

save(db_proc92, file = here("input/data/proc/db_proc92.RData"))
```



# ISSP 1999

## Select

```{r}

db_proc99 <- issp99 %>% 
  select(year = v1, 
         v3,
         sex,
         age,
         educyrs,
         degree,
         rincome,
         rincomer,
         ess = v46,
         nmerit_wealthy_fam = v4,
         nmerit_contacts = v5,
         merit_effort = v6,
         merit_talent = v7,
         just_health = v39,
         just_educ = v40,
         perc_inequality = v57,
         isco88_4,
         selfemp,
         nemploy,
         wrksup,
         nwrksup,
         weight) %>% 
  janitor::clean_names()

```


##  Recode and transform

### Year

```{r}
frq(db_proc99$year)
db_proc99$year <- as.numeric(db_proc99$year)
frq(db_proc99$year)
```

### Country

```{r}
frq(db_proc99$v3)
db_proc99 <- db_proc99 %>% 
  mutate(country = case_when(
    v3 == 1  ~ "Australia",
    v3 == 2 | v3 == 3  ~ "Germany",
    v3 == 4  ~ "United Kingdom",
    v3 == 5  ~ "Ireland",
    v3 == 6  ~ "United States",
    v3 == 7  ~ "Austria",
    v3 == 8  ~ "Hungary",
    v3 == 10 ~ "Ireland",      
    v3 == 12 ~ "Norway",
    v3 == 13 ~ "Sweden",
    v3 == 14 ~ "Czechia",  
    v3 == 15 ~ "Slovenia",
    v3 == 16 ~ "Poland",
    v3 == 17 ~ "Bulgaria",
    v3 == 18 ~ "Russia",
    v3 == 19 ~ "New Zealand",
    v3 == 20 ~ "Canada",
    v3 == 21 ~ "Philippines",
    v3 == 22 ~ "Israel",
    v3 == 24 ~ "Japan",
    v3 == 25 ~ "Spain",
    v3 == 26 ~ "Latvia",
    v3 == 27 ~ "France",
    v3 == 28 ~ "Cyprus",
    v3 == 29 ~ "Portugal",
    v3 == 30 ~ "Chile",
    v3 == 33 ~ "Slovakia",
    TRUE ~ NA_character_
  ))
db_proc99$iso_country <- countrycode::countrycode(db_proc99$country, origin = "country.name", destination = "iso3c")
frq(db_proc99$country); frq(db_proc99$iso_country)

```

### Sex
```{r}
frq(db_proc99$sex)
db_proc99$sex <- set_na(db_proc99$sex, na = c(9), drop.levels = F)
db_proc99 <- db_proc99 %>% 
mutate(sex = if_else(sex == 1, 'Male', 'Female'),
       sex = factor(sex, levels = c("Male", "Female")))
frq(db_proc99$sex)
```


### Age
```{r}
summary(db_proc99$age)
db_proc99$age <- set_na(db_proc99$age, na = c(98,99), drop.levels = F)
db_proc99$age <- as.numeric(db_proc99$age)

```

### Education
```{r}
# Eucation
frq(db_proc99$educyrs) # We don't know if this is equal in every country; we use degree
frq(db_proc99$degree)
db_proc99$degree <- set_na(db_proc99$degree, na = c(8,9), drop.levels = F)
db_proc99 <- db_proc99 %>% 
  mutate(educ = case_when(degree %in% c(1,2) ~ "Incomplete primary or less",
                            degree == 3 ~ "Complete primary",
                            degree == 4 ~ "Incomplete secondary",
                            degree == 5 ~ "Complete secondary",
                            degree == 6 ~ "Incomplete university",
                            degree == 7 ~ "Complete university",
                            TRUE ~ NA_character_),
           educ = factor(educ, 
                           levels = c("Incomplete primary or less",
                           "Complete primary",
                           "Incomplete secondary",
                           "Complete secondary",
                           "Incomplete university",
                           "Complete university")))

db_proc99 <- db_proc99 %>% 
  mutate(educ_dic = case_when(degree %in% c(1:6) ~ 'Less than universitary',
                              degree == 7 ~ 'Universitary',
                              TRUE ~ NA_character_),
         educ_dic = factor(educ_dic, levels = c("Universitary", "Less than universitary")))
frq(db_proc99$educ)
frq(db_proc99$educ_dic)
```


### Income
```{r}

# Income continuous
summary(db_proc99$rincome)
db_proc99 <- db_proc99 %>%
  mutate(
    rincome = haven::zap_labels(rincome),
    rincome = sjmisc::set_na(rincome, na = c(999997,999998,999999))
  )

# ntile function without NAs
ntile_na <- function(x,n)
{
  notna <- !is.na(x)
  out <- rep(NA_real_,length(x))
  out[notna] <- ntile(x[notna],n)
  return(out)
}

db_proc99 %>% 
  mutate(decil = ntile_na(rincome, 10)) %>% 
  frq(decil)

# Income decil
frq(db_proc99$rincomer) 
db_proc99$rincomer <- set_na(db_proc99$rincomer, na = c(97,98,99), drop.levels = F) # there is a decil 0; those no income no labor force
db_proc99$income_decil <- as.numeric(db_proc99$rincomer)
frq(db_proc99$income_decil) 
```

### Subjective social status
```{r}
frq(db_proc99$ess)
db_proc99$ess <- set_na(db_proc99$ess, na = c(97,98,99), drop.levels = F)
db_proc99$ess <- as.numeric(db_proc99$ess)
frq(db_proc99$ess)
```

### Meritocratic variables
```{r}
frq(db_proc99$merit_effort)
frq(db_proc99$merit_talent)

db_proc99 <- db_proc99 %>% 
  mutate(
    across(
      .cols = c(merit_effort, merit_talent),
      .fns = ~ set_na(., na = c(8,9), drop.levels = F)
    )
  ) # ok

db_proc99$merit_effort <- sjmisc::rec(db_proc99$merit_effort,rec = "rev")
db_proc99$merit_talent <- sjmisc::rec(db_proc99$merit_talent,rec = "rev")

frq(db_proc99$merit_effort)
frq(db_proc99$merit_talent)
```

### Non-meritocratic variables
```{r}
frq(db_proc99$nmerit_wealthy_fam)
frq(db_proc99$nmerit_contacts)

db_proc99 <- db_proc99 %>% 
  mutate(
    across(
      .cols = c(nmerit_wealthy_fam, nmerit_contacts),
      .fns = ~ set_na(., na = c(8,9), drop.levels = F)
    )
  ) # ok

db_proc99$nmerit_wealthy_fam <- sjmisc::rec(db_proc99$nmerit_wealthy_fam,rec = "rev")
db_proc99$nmerit_contacts <- sjmisc::rec(db_proc99$nmerit_contacts,rec = "rev")

frq(db_proc99$nmerit_wealthy_fam)
frq(db_proc99$nmerit_contacts)
```

### Market justice healthcare
```{r}
frq(db_proc99$just_health)
db_proc99$just_health <- set_na(db_proc99$just_health, na = c(8,9), drop.levels = F) 
db_proc99$just_health <- sjmisc::rec(db_proc99$just_health,rec = "rev")
frq(db_proc99$just_health)
?rec
```

### Market justice in education
```{r}
frq(db_proc99$just_educ)
db_proc99$just_educ <- set_na(db_proc99$just_educ, na = c(8,9), drop.levels = F) 
db_proc99$just_educ <- sjmisc::rec(db_proc99$just_educ,rec = "rev")
frq(db_proc99$just_educ)

```

### Perceived inequality
```{r}
frq(db_proc99$perc_inequality)
db_proc99$perc_inequality <- set_na(db_proc99$perc_inequality, na = c(8,9), drop.levels = F)
db_proc99$perc_inequality <- sjmisc::rec(db_proc99$perc_inequality,rec = "rev")
frq(db_proc99$perc_inequality)
```

### Class

```{r}
db_proc99 <- db_proc99 %>%
  mutate(
    selfemp_bin = if_else(selfemp == 1, 1, 0),
    nemploy_rec = case_when(
      nemploy == 9995 ~ 0,
      nemploy %in% c(9997, 9999) ~ NA_real_,
      TRUE ~ nemploy
    ),
    egp5 = isco88toEGP(isco88_4, n.employees = nemploy_rec,
                        self.employed = selfemp_bin, n.classes = 5)
  )

frq(db_proc99$egp5)
```

## Save

```{r}
db_proc99 <- db_proc99 %>% 
  select(year, country, iso_country, sex, age, educ, educ_dic, income_decil, ess, just_health, just_educ, perc_inequality, egp5,  merit_effort, merit_talent, nmerit_wealthy_fam, nmerit_contacts, weight)

save(db_proc99, file = here("input/data/proc/db_proc99.RData"))
```

# Processing ISSP 2009

## Select

```{r}
db_proc09 <- issp09 %>% 
  janitor::clean_names() %>% 
  select(v5, 
         sex,
         age,
         degree,
         141:181,
         ess = v44,
         nmerit_wealthy_fam = v6, 
         nmerit_educated_parents = v7, 
         merit_ambition = v9, 
         merit_effort = v10,
         nmerit_contacts = v11,
         nmerit_political_connection = v12,
         nmerit_race = v14,
         nmerit_gender = v16,
         just_health = v38,
         just_educ = v39,
         perc_inequality = v54,
         isco88,
         wrktype,
         nemploy,
         wrksup,
         weight
         )

```


##  Recode and transform

### Year

```{r}
db_proc09$year <- 2009
db_proc09$year <- as.numeric(db_proc09$year)
```

### Country

```{r}
frq(db_proc09$v5)
db_proc09$country <- NA
db_proc09$country <- countrycode::countrycode(db_proc09$v5, 
                                              origin = "iso3n", 
                                              destination = "country.name")
db_proc09$iso_country <- countrycode::countrycode(db_proc09$country, 
                                                  origin = "country.name",
                                                  destination = "iso3c")
frq(db_proc09$country); frq(db_proc09$iso_country)

```

### Sex
```{r}
frq(db_proc09$sex)
db_proc09$sex <- set_na(db_proc09$sex, na = c(9), drop.levels = F)
db_proc09 <- db_proc09 %>% 
  mutate(sex = if_else(sex == 1, 'Male', 'Female'),
       sex = factor(sex, levels = c("Male", "Female")))
frq(db_proc09$sex)
```


### Age
```{r}
summary(db_proc09$age)
db_proc09$age <- set_na(db_proc09$age, na = c(99), drop.levels = F)
db_proc09$age <- as.numeric(db_proc09$age)
```

### Education
```{r}
frq(db_proc09$degree)
db_proc09$degree <- set_na(db_proc09$degree, na = c(8,9), drop.levels = F)
db_proc09 <- db_proc09 %>% 
  mutate(educ = case_when(degree == 0 ~ "Incomplete primary or less",
                            degree == 1 ~ "Complete primary",
                            degree == 2 ~ "Incomplete secondary",
                            degree == 3 ~ "Complete secondary",
                            degree == 4 ~ "Incomplete university",
                            degree == 5 ~ "Complete university",
                            TRUE ~ NA_character_),
           educ = factor(educ, 
                           levels = c("Incomplete primary or less",
                           "Complete primary",
                           "Incomplete secondary",
                           "Complete secondary",
                           "Incomplete university",
                           "Complete university")))


db_proc09 <- db_proc09 %>% 
  mutate(educ_dic = case_when(degree %in% c(0:4) ~ 'Less than universitary',
                              degree == 5 ~ 'Universitary',
                              TRUE ~ NA_character_),
         educ_dic = factor(educ_dic, levels = c("Universitary", "Less than universitary")))
frq(db_proc09$educ)
frq(db_proc09$educ_dic)
```


### Income
```{r}
#frq(db_proc09$ar_rinc)
#frq(db_proc09$at_rinc)
#frq(db_proc09$au_rinc)
#frq(db_proc09$be_rinc)
#frq(db_proc09$bg_rinc)
#frq(db_proc09$ch_rinc)
#frq(db_proc09$cl_rinc)
#frq(db_proc09$cn_rinc)
#frq(db_proc09$cy_rinc)
#frq(db_proc09$cz_rinc)
#frq(db_proc09$de_rinc)
#frq(db_proc09$dk_rinc)
#frq(db_proc09$ee_rinc)
#frq(db_proc09$es_rinc)
#frq(db_proc09$fi_rinc)
#frq(db_proc09$fr_rinc)
#frq(db_proc09$gb_rinc)
#frq(db_proc09$hr_rinc)
#frq(db_proc09$hu_rinc)
#frq(db_proc09$il_rinc)
#frq(db_proc09$is_rinc)
#frq(db_proc09$it_rinc)
#frq(db_proc09$jp_rinc)
#frq(db_proc09$kr_rinc)
#frq(db_proc09$lt_rinc)
#frq(db_proc09$lv_rinc)
#frq(db_proc09$no_rinc)
#frq(db_proc09$nz_rinc)
#frq(db_proc09$ph_rinc)
#frq(db_proc09$pl_rinc)
#frq(db_proc09$pt_rinc)
#frq(db_proc09$ru_rinc)
#frq(db_proc09$se_rinc)
#frq(db_proc09$si_rinc)
#frq(db_proc09$sk_rinc)
#frq(db_proc09$tr_rinc)
#frq(db_proc09$tw_rinc)
#frq(db_proc09$ua_rinc)
#frq(db_proc09$us_rinc)
#frq(db_proc09$ve_rinc)
#frq(db_proc09$za_rinc)

# ntile function without NAs
ntile_zero_na <- function(x, n = 10) {
  out   <- rep(NA_real_, length(x))
  is_na <- is.na(x)
  is0   <- !is_na & x == 0
  pos   <- !is_na & x > 0
  out[is0] <- 0
  if (any(pos)) out[pos] <- ntile(x[pos], n)
  out
}

db_proc09 <- db_proc09 %>% 
 mutate_at(vars(6:46), ~ as.numeric(.)) %>% 
  mutate_at(vars(6:46), funs(car::recode(. ,"999990 = NA; 999997 = NA; 999998 = NA; 999999 = NA; 
                                              9999990 = NA; 9999997 = NA; 9999998 = NA; 9999999 = NA;
                                              99999990 = NA; 99999997 = NA; 99999998 = NA; 99999999 = NA"))) %>% 
  mutate(
    income_personal = do.call(coalesce, across(6:46)),
    income_decil    = ntile_zero_na(income_personal, 10)
  )

db_proc09 %>% 
  mutate(n_non_na = rowSums(!is.na(pick(6:46)))) %>% 
  count(n_non_na)
frq(db_proc09$income_decil)

```

### Subjective social status
```{r}
frq(db_proc09$ess)
db_proc09$ess <- set_na(db_proc09$ess, na = c(97,98,99), drop.levels = F)
db_proc09$ess <- as.numeric(db_proc09$ess)
frq(db_proc09$ess)
```

### Meritocratic variables
```{r}
frq(db_proc09$merit_effort)
frq(db_proc09$merit_ambition)

db_proc09 <- db_proc09 %>% 
  mutate(
    across(
      .cols = c(merit_effort, merit_ambition),
      .fns = ~ set_na(., na = c(8,9), drop.levels = F)
    )
  ) # ok

db_proc09$merit_effort <- sjmisc::rec(db_proc09$merit_effort,rec = "rev")
db_proc09$merit_ambition <- sjmisc::rec(db_proc09$merit_ambition,rec = "rev")

frq(db_proc09$merit_effort)
frq(db_proc09$merit_ambition)
```

### Non-meritocratic variables
```{r}
frq(db_proc09$nmerit_wealthy_fam)
frq(db_proc09$nmerit_contacts)
frq(db_proc09$nmerit_educated_parents)
frq(db_proc09$nmerit_political_connection)
frq(db_proc09$nmerit_race)
frq(db_proc09$nmerit_gender)

db_proc09 <- db_proc09 %>% 
  mutate(
    across(
      .cols = c(nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents,
                nmerit_political_connection, nmerit_race, nmerit_gender),
      .fns = ~ set_na(., na = c(8,9), drop.levels = F)
    )
  ) # ok

db_proc09$nmerit_wealthy_fam <- sjmisc::rec(db_proc09$nmerit_wealthy_fam,rec = "rev")
db_proc09$nmerit_contacts <- sjmisc::rec(db_proc09$nmerit_contacts,rec = "rev")
db_proc09$nmerit_educated_parents <- sjmisc::rec(db_proc09$nmerit_educated_parents,rec = "rev")
db_proc09$nmerit_political_connection <- sjmisc::rec(db_proc09$nmerit_political_connection,rec = "rev")
db_proc09$nmerit_race <- sjmisc::rec(db_proc09$nmerit_race,rec = "rev")
db_proc09$nmerit_gender <- sjmisc::rec(db_proc09$nmerit_gender,rec = "rev")

frq(db_proc09$nmerit_wealthy_fam)
frq(db_proc09$nmerit_contacts)
frq(db_proc09$nmerit_educated_parents)
frq(db_proc09$nmerit_political_connection)
frq(db_proc09$nmerit_race)
frq(db_proc09$nmerit_gender)

```

### Market justice healthcare
```{r}
frq(db_proc09$just_health)
db_proc09$just_health <- set_na(db_proc09$just_health, na = c(8,9), drop.levels = F) 
db_proc09$just_health <- sjmisc::rec(db_proc09$just_health,rec = "rev")
frq(db_proc09$just_health)

```

### Market justice in education
```{r}
frq(db_proc09$just_educ)
db_proc09$just_educ <- set_na(db_proc09$just_educ, na = c(8,9), drop.levels = F) 
db_proc09$just_educ <- sjmisc::rec(db_proc09$just_educ,rec = "rev")
frq(db_proc09$just_educ)

```

### Perceived inequality
```{r}
frq(db_proc09$perc_inequality)
db_proc09$perc_inequality <- set_na(db_proc09$perc_inequality, na = c(8,9), drop.levels = F)
db_proc09$perc_inequality <- sjmisc::rec(db_proc09$perc_inequality,rec = "rev")
frq(db_proc09$perc_inequality)
```

### Class

```{r}

db_proc09 <- db_proc09 %>% 
  mutate(
    selfemp_bin = if_else(wrktype == 4, 1, 0),
    nemploy_rec = case_when(
      nemploy == 9995 ~ 0,
      nemploy %in% c(9997, 9998, 9999) ~ NA_real_,
      TRUE ~ nemploy
    ),
    egp5  = isco88toEGP(isco88, n.employees = nemploy_rec,
                         self.employed = selfemp_bin, n.classes = 5)
  )

frq(db_proc09$egp5)
```

## Save

```{r}
#| collapse: false

db_proc09 <- db_proc09 %>% 
  select(year, country, iso_country, sex, age, educ, educ_dic, income_decil, ess, just_health, just_educ, perc_inequality, egp5, merit_effort, merit_ambition, nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents, nmerit_political_connection, nmerit_race, nmerit_gender, weight)
```


```{r}
save(db_proc09, file = here("input/data/proc/db_proc09.RData"))
```

# Processing ISSP 2019

## Select

```{r}
db_proc19 <- issp19 %>% 
  janitor::clean_names() %>% 
  select(country, 
         sex,
         age,
         degree,
         258:286,
         ess = topbot,
         nmerit_wealthy_fam = v1, 
         nmerit_educated_parents = v2, 
         merit_effort = v4,
         nmerit_contacts = v5,
         nmerit_political_connection = v6,
         nmerit_race = v8,
         nmerit_gender = v10,
         just_health = v30,
         just_educ = v31,
         perc_inequality = v48,
         isco08,
         emprel,
         wrksup,
         nsup,
         weight
         )
```


##  Recode and transform

### Year

```{r}
db_proc19$year <- 2019
db_proc19$year <- as.numeric(db_proc19$year)
```

### Country

```{r}
frq(db_proc19$country)
db_proc19$country <- countrycode::countrycode(db_proc19$country, 
                                              origin = "iso3n", 
                                              destination = "country.name")
db_proc19$iso_country <- countrycode::countrycode(db_proc19$country, 
                                                  origin = "country.name",
                                                  destination = "iso3c")
frq(db_proc19$country); frq(db_proc19$iso_country)

```

### Sex
```{r}
frq(db_proc19$sex)
db_proc19$sex <- set_na(db_proc19$sex, na = c(-9), drop.levels = F)
db_proc19 <- db_proc19 %>% 
  mutate(sex = if_else(sex == 1, 'Male', 'Female'),
       sex = factor(sex, levels = c("Male", "Female")))
frq(db_proc19$sex)
```


### Age
```{r}
summary(db_proc19$age)
db_proc19$age <- set_na(db_proc19$age, na = c(-9), drop.levels = F)
db_proc19$age <- as.numeric(db_proc19$age)
```

### Education
```{r}
frq(db_proc19$degree)
db_proc19$degree <- set_na(db_proc19$degree, na = c(-9), drop.levels = F)
db_proc19 <- db_proc19 %>% 
  mutate(educ = case_when(degree == 0 ~ "Incomplete primary or less",
                            degree == 1 ~ "Complete primary",
                            degree == 2 ~ "Incomplete secondary",
                            degree == 3 ~ "Complete secondary",
                            degree == 4 ~ "Incomplete university",
                            degree %in% c(5,6) ~ "Complete university",
                            TRUE ~ NA_character_),
           educ = factor(educ, 
                           levels = c("Incomplete primary or less",
                           "Complete primary",
                           "Incomplete secondary",
                           "Complete secondary",
                           "Incomplete university",
                           "Complete university")))

db_proc19 <- db_proc19 %>% 
  mutate(educ_dic = case_when(degree %in% c(0:4) ~ 'Less than universitary',
                              degree %in% c(5:6) ~ 'Universitary',
                              TRUE ~ NA_character_),
         educ_dic = factor(educ_dic, levels = c("Universitary", "Less than universitary")))
frq(db_proc19$educ)
frq(db_proc19$educ_dic)
```


### Income
```{r}
#frq(db_proc19$at_rinc)
#frq(db_proc19$au_rinc)
#frq(db_proc19$bg_rinc)
#frq(db_proc19$ch_rinc)
#frq(db_proc19$cl_rinc)
#frq(db_proc19$cz_rinc)
#frq(db_proc19$de_rinc)
#frq(db_proc19$dk_rinc)
#frq(db_proc19$fi_rinc)
#frq(db_proc19$fr_rinc)
#frq(db_proc19$gb_rinc)
#frq(db_proc19$hr_rinc)
#frq(db_proc19$il_rinc)
#frq(db_proc19$is_rinc)
#frq(db_proc19$it_rinc)
#frq(db_proc19$jp_rinc)
#frq(db_proc19$lt_rinc)
#frq(db_proc19$no_rinc)
#frq(db_proc19$nz_rinc)
#frq(db_proc19$ph_rinc)
#frq(db_proc19$ru_rinc)
#frq(db_proc19$se_rinc)
#frq(db_proc19$si_rinc)
#frq(db_proc19$sr_rinc)
#frq(db_proc19$th_rinc)
#frq(db_proc19$tw_rinc)
#frq(db_proc19$us_rinc)
#frq(db_proc19$ve_rinc)
#frq(db_proc19$za_rinc)

# ntile function without NAs
ntile_zero_na <- function(x, n = 10) {
  out   <- rep(NA_real_, length(x))
  is_na <- is.na(x)
  is0   <- !is_na & x == 0
  pos   <- !is_na & x > 0
  out[is0] <- 0
  if (any(pos)) out[pos] <- ntile(x[pos], n)
  out
}

db_proc19 <- db_proc19 %>% 
  mutate_at(vars(5:34), ~ as.numeric(.)) %>% 
 mutate_at(vars(5:34), funs(car::recode(. ,"-9 = NA; -8 = NA; -7 = NA; -2 = NA"))) %>% 
  mutate(
    income_personal = do.call(coalesce, across(5:34)),
    income_decil    = ntile_zero_na(income_personal, 10)
  )

db_proc19 %>% 
  mutate(n_non_na = rowSums(!is.na(pick(5:34)))) %>% 
  count(n_non_na)
frq(db_proc19$income_decil)
```

### Subjective social status
```{r}
frq(db_proc19$ess)
db_proc19$ess <- set_na(db_proc19$ess, na = c(-8,-9), drop.levels = F)
db_proc19$ess <- as.numeric(db_proc19$ess)
frq(db_proc19$ess)
```

### Meritocratic variables
```{r}
frq(db_proc19$merit_effort)

db_proc19 <- db_proc19 %>% 
  mutate(
    across(
      .cols = c(merit_effort),
      .fns = ~ set_na(., na = c(-8,-9), drop.levels = F)
    )
  ) # ok

db_proc19$merit_effort <- sjmisc::rec(db_proc19$merit_effort,rec = "rev")

frq(db_proc19$merit_effort)
```

### Non-meritocratic variables
```{r}
frq(db_proc19$nmerit_wealthy_fam)
frq(db_proc19$nmerit_contacts)
frq(db_proc19$nmerit_educated_parents)
frq(db_proc19$nmerit_political_connection)
frq(db_proc19$nmerit_race)
frq(db_proc19$nmerit_gender)

db_proc19 <- db_proc19 %>% 
  mutate(
    across(
      .cols = c(nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents,
                nmerit_political_connection, nmerit_race, nmerit_gender),
      .fns = ~ set_na(., na = c(-8,-9), drop.levels = F)
    )
  ) # ok

db_proc19$nmerit_wealthy_fam <- sjmisc::rec(db_proc19$nmerit_wealthy_fam,rec = "rev")
db_proc19$nmerit_contacts <- sjmisc::rec(db_proc19$nmerit_contacts,rec = "rev")
db_proc19$nmerit_educated_parents <- sjmisc::rec(db_proc19$nmerit_educated_parents,rec = "rev")
db_proc19$nmerit_political_connection <- sjmisc::rec(db_proc19$nmerit_political_connection,rec = "rev")
db_proc19$nmerit_race <- sjmisc::rec(db_proc19$nmerit_race,rec = "rev")
db_proc19$nmerit_gender <- sjmisc::rec(db_proc19$nmerit_gender,rec = "rev")

frq(db_proc19$nmerit_wealthy_fam)
frq(db_proc19$nmerit_contacts)
frq(db_proc19$nmerit_educated_parents)
frq(db_proc19$nmerit_political_connection)
frq(db_proc19$nmerit_race)
frq(db_proc19$nmerit_gender)

```

### Market justice healthcare
```{r}
frq(db_proc19$just_health)
db_proc19$just_health <- set_na(db_proc19$just_health, na = c(-8, -9), drop.levels = F) 
db_proc19$just_health <- sjmisc::rec(db_proc19$just_health,rec = "rev")
frq(db_proc19$just_health)

```

### Market justice in education
```{r}
frq(db_proc19$just_educ)
db_proc19$just_educ <- set_na(db_proc19$just_educ, na = c(-8,-9), drop.levels = F) 
db_proc19$just_educ <- sjmisc::rec(db_proc19$just_educ,rec = "rev")
frq(db_proc19$just_educ)

```

### Perceived inequality
```{r}
frq(db_proc19$perc_inequality)
db_proc19$perc_inequality <- set_na(db_proc19$perc_inequality, na = c(-8,-9,-4), drop.levels = F)
db_proc19$perc_inequality <- sjmisc::rec(db_proc19$perc_inequality,rec = "rev")
frq(db_proc19$perc_inequality)
```

### Class

```{r}
db_proc19 <- db_proc19 %>%
  mutate(
    isco08 = set_na(isco08, na = c(-9,-8,-4), drop.levels = F),
    emprel = set_na(emprel, na = c(-9,-4), drop.levels = F),
    isco88_4 = isco08to88(isco08),
    selfemp_bin = if_else(emprel %in% c(2,3,4), 1, 0),
    nemploy_rec = case_when(
      emprel == 2         ~ 0,   # self-employed sin empleados
      emprel %in% c(3,4)  ~ 1,   # self-employed con empleados
      TRUE                 ~ NA_real_
    ),
    egp5  = isco88toEGP(isco88_4, n.employees = nemploy_rec,
                         self.employed = selfemp_bin, n.classes = 5)
  )

frq(db_proc19$egp5)
```

## Save

```{r}
#| collapse: false

db_proc19 <- db_proc19 %>%
  select(year, country, iso_country, sex, age, educ, educ_dic, income_decil, ess, just_health, just_educ, perc_inequality, merit_effort, nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents, nmerit_political_connection, nmerit_race, nmerit_gender, egp5, weight)


save(db_proc19, file = here("input/data/proc/db_proc19.RData"))
```

# Final data

## Merge ISSP

```{r}
#| warning: false
#| message: false
dfs_no_labels <- map(list(db_proc87, db_proc92, db_proc99, db_proc09, db_proc19), remove_all_labels)

names(dfs_no_labels) <- c("db_proc87", "db_proc92", "db_proc99", "db_proc09", "db_proc19")
list2env(dfs_no_labels, envir = .GlobalEnv)

db_proc <- bind_rows(
  `1987` = db_proc87,
  `1992` = db_proc92,
  `1999` = db_proc99,
  `2009` = db_proc09,
  `2019` = db_proc19,
  .id = "source_year"
) %>%
  select(-source_year)

db_proc <- tibble::rowid_to_column(db_proc, "id_subject")

head(db_proc)
```

## Save

```{r}
save(db_proc, file = here("input/data/proc/issp_proc.RData"))
```

# Cumulative data


## Select

```{r}

db_cumu <- issp_cumu %>%
  select(year, country,
         sex = SEX,
         age = AGE,
         degree = DEGREE,
         income = INCOME,
         ess = TOPBOT,
         merit_effort = v6,
         nmerit_contacts = v7,
         nmerit_race = v9,
         nmerit_gender = v12,
         just_health = v37,
         just_educ = v38,
         perc_inequality = v54,
         isco88 = ISCO88,
         wrksup = WRKSUP,
         selfemp = SELFEMP,
         weight = WEIGHT) %>%
  janitor::clean_names()

```



```{r}
na_codes_cumu <- c(-9,-8,-7,-4,-3,-1)
```

##  Recode and transform

### Year

```{r}
frq(db_cumu$year)
db_cumu$year <- as.numeric(db_cumu$year)
```

### Country

```{r}
db_cumu$country <- as.numeric(db_cumu$country)
db_cumu$country <- countrycode::countrycode(db_cumu$country,
                                             origin = "iso3n",
                                             destination = "country.name")
db_cumu$iso_country <- countrycode::countrycode(db_cumu$country,
                                                 origin = "country.name",
                                                 destination = "iso3c")
frq(db_cumu$country); frq(db_cumu$iso_country)

```

### Sex
```{r}
frq(db_cumu$sex)
db_cumu$sex <- set_na(db_cumu$sex, na = na_codes_cumu, drop.levels = F)
db_cumu <- db_cumu %>%
  mutate(sex = if_else(sex == 1, 'Male', 'Female'),
         sex = factor(sex, levels = c("Male", "Female")))
frq(db_cumu$sex)
```

### Age
```{r}
summary(db_cumu$age)
db_cumu$age <- set_na(db_cumu$age, na = na_codes_cumu, drop.levels = F)
db_cumu$age <- as.numeric(db_cumu$age)
```

### Education
```{r}
frq(db_cumu$degree)
db_cumu$degree <- set_na(db_cumu$degree, na = na_codes_cumu, drop.levels = F)
db_cumu <- db_cumu %>%
  mutate(educ = case_when(degree == 0 ~ "Incomplete primary or less",
                            degree == 1 ~ "Complete primary",
                            degree == 2 ~ "Incomplete secondary",
                            degree == 3 ~ "Complete secondary",
                            degree == 4 ~ "Incomplete university",
                            degree == 5 ~ "Complete university",
                            TRUE ~ NA_character_),
           educ = factor(educ,
                           levels = c("Incomplete primary or less",
                           "Complete primary",
                           "Incomplete secondary",
                           "Complete secondary",
                           "Incomplete university",
                           "Complete university")))

db_cumu <- db_cumu %>%
  mutate(educ_dic = case_when(degree %in% c(0:4) ~ 'Less than universitary',
                              degree == 5 ~ 'Universitary',
                              TRUE ~ NA_character_),
         educ_dic = factor(educ_dic, levels = c("Universitary", "Less than universitary")))
frq(db_cumu$educ)
frq(db_cumu$educ_dic)
```

### Income
```{r}
frq(db_cumu$income)
db_cumu$income <- set_na(db_cumu$income, na = na_codes_cumu, drop.levels = F)
frq(db_cumu$income)
```


### Subjective social status
```{r}
frq(db_cumu$ess)
db_cumu$ess <- set_na(db_cumu$ess, na = na_codes_cumu, drop.levels = F)
db_cumu$ess <- as.numeric(db_cumu$ess)
frq(db_cumu$ess)
```

### Meritocratic variables
```{r}
frq(db_cumu$merit_effort)

db_cumu <- db_cumu %>%
  mutate(
    across(
      .cols = c(merit_effort),
      .fns = ~ set_na(., na = na_codes_cumu, drop.levels = F)
    )
  ) # ok

db_cumu$merit_effort <- sjmisc::rec(db_cumu$merit_effort,rec = "rev")

frq(db_cumu$merit_effort)
```

### Non-meritocratic variables
```{r}
frq(db_cumu$nmerit_contacts)
frq(db_cumu$nmerit_race)
frq(db_cumu$nmerit_gender)

db_cumu <- db_cumu %>%
  mutate(
    across(
      .cols = c(nmerit_contacts, nmerit_race, nmerit_gender),
      .fns = ~ set_na(., na = na_codes_cumu, drop.levels = F)
    )
  ) # ok

db_cumu$nmerit_contacts <- sjmisc::rec(db_cumu$nmerit_contacts,rec = "rev")
db_cumu$nmerit_race <- sjmisc::rec(db_cumu$nmerit_race,rec = "rev")
db_cumu$nmerit_gender <- sjmisc::rec(db_cumu$nmerit_gender,rec = "rev")

frq(db_cumu$nmerit_contacts)
frq(db_cumu$nmerit_race)
frq(db_cumu$nmerit_gender)

```

### Market justice healthcare
```{r}
frq(db_cumu$just_health)
db_cumu$just_health <- set_na(db_cumu$just_health, na = na_codes_cumu, drop.levels = F)
db_cumu$just_health <- sjmisc::rec(db_cumu$just_health,rec = "rev")
frq(db_cumu$just_health)
```

### Market justice in education
```{r}
frq(db_cumu$just_educ)
db_cumu$just_educ <- set_na(db_cumu$just_educ, na = na_codes_cumu, drop.levels = F)
db_cumu$just_educ <- sjmisc::rec(db_cumu$just_educ,rec = "rev")
frq(db_cumu$just_educ)
```

### Perceived inequality
```{r}
frq(db_cumu$perc_inequality)
db_cumu$perc_inequality <- set_na(db_cumu$perc_inequality, na = na_codes_cumu, drop.levels = F)
db_cumu$perc_inequality <- sjmisc::rec(db_cumu$perc_inequality,rec = "rev")
frq(db_cumu$perc_inequality)
```

### Class

```{r}
db_cumu <- db_cumu %>%
  mutate(
    isco88 = set_na(isco88, na = na_codes_cumu, drop.levels = F),
    wrksup = set_na(wrksup, na = na_codes_cumu, drop.levels = F),
    selfemp = set_na(selfemp, na = na_codes_cumu, drop.levels = F),
    selfemp_bin = if_else(selfemp == 1, 1, 0),
    nemploy = NA_real_, # no existe una variable de número de empleados armonizada en el cumulativo
    egp5 = isco88toEGP(isco88, n.employees = nemploy,
                        self.employed = selfemp_bin, n.classes = 5)
  )

frq(db_cumu$egp5)
```

## Save

```{r}
db_cumu <- db_cumu %>%
  select(year, country, iso_country, sex, age, educ, educ_dic, income, ess, just_health, just_educ, perc_inequality, egp5, merit_effort, nmerit_contacts, nmerit_race, nmerit_gender, isco88, wrksup, nemploy, weight)

save(db_cumu, file = here("input/data/proc/db_cumu.RData"))
```

# Assessment of the availability of variables and countries

## Variables available per wave

```{r}
#| label: tbl-vars-cumu
#| tbl-cap: "Missing data by variable and wave (% NA)"
#| tbl-cap-location: top
#| results: asis
#| warning: false

tab_avail_cumu <- db_cumu %>%
  group_by(year) %>%
  summarise(
    n = n(),
    just_health = mean(is.na(just_health)) * 100,
    just_educ = mean(is.na(just_educ)) * 100,
    perc_inequality = mean(is.na(perc_inequality)) * 100,
    merit_effort = mean(is.na(merit_effort)) * 100,
    ess = mean(is.na(ess)) * 100,
    isco88 = mean(is.na(isco88)) * 100,
    wrksup = mean(is.na(wrksup)) * 100,
    .groups = "drop"
  ) %>%
  mutate(across(-c(year, n), ~ round(., 1)),
         year = factor(year))

num_cols_cumu <- setdiff(names(tab_avail_cumu), c("year", "n"))

tab_avail_cumu_fmt <- tab_avail_cumu %>%
  mutate(across(
    all_of(num_cols_cumu),
    ~ ifelse(
      is.na(.x), "",
      ifelse(
        .x == 100,
        kableExtra::cell_spec(sprintf("%.1f", .x), "html", color = "white", background = "darkred"),
        sprintf("%.1f", .x)
      )
    )
  ))

kableExtra::kable(tab_avail_cumu_fmt,
                  format = "html",
                  booktabs = T,
                  escape = F,
                  align = 'c',
                  format.args = list(big.mark = ",")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", full_width = FALSE))

```


## Countries per wave


```{r}
#| label: tbl-country-wave-avail
#| tbl-cap: "Available countries by wave"
#| tbl-cap-location: top
#| results: asis
#| warning: false

tab_country_wave <- db_proc %>%
  distinct(country, year) %>%
  mutate(available = 1L) %>%
  tidyr::pivot_wider(names_from = year, values_from = available, values_fill = 0L) %>%
  arrange(country)

wave_cols <- setdiff(names(tab_country_wave), "country")
tab_country_wave <- tab_country_wave %>%
  select(country, all_of(sort(wave_cols))) %>%
  mutate(Total = rowSums(across(all_of(wave_cols)))) %>%
  janitor::adorn_totals("row")

tab_country_wave_fmt <- tab_country_wave %>%
  mutate(across(
    all_of(wave_cols),
    ~ ifelse(
      .x == 1,
      kableExtra::cell_spec(.x, "html", color = "white", background = "darkred"),
      as.character(.x)
    )
  ))

kableExtra::kable(tab_country_wave_fmt,
                  format = "html",
                  booktabs = T,
                  escape = F,
                  align = 'c') %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", full_width = FALSE))

```

## Suggestions for Analysis Criteria

- Due to the absence of market justice variables in the 1987 and 1992 waves, we have retained the **last 3 ISSP waves (1999, 2009, 2019)**

- By using the last 3 waves, we have **96 country-waves** in the total sample

- If we apply the selection criterion of including only countries that appear in **at least two** versions of the survey, the number of country-waves decreases to **86** (Argentina, Belgium, China, Estonia, Ireland, South Korea, Suriname, Thailand, Turkey, and Ukraine are excluded)

# Descriptives

## Main variables per country and wave

```{r}
#| label: tbl-vars
#| tbl-cap: "Missing values percent per variable by country-wave"
#| tbl-cap-location: top
#| results: asis
#| echo: false
#| warning: false

db_proc$wave <- as.factor(db_proc$year)
db_proc$country_wave <- do.call(paste, c(db_proc[c("iso_country", "wave")], sep = "_"))

tab <- db_proc %>%
  select(country_wave, just_health,  just_educ, ess, egp5, starts_with("merit"), starts_with("nmerit")) %>%
  group_by(country_wave) %>%
  summarise(
    across(
      everything(),                     
      list(
        n_complete = ~sum(!is.na(.)),
        n_na       = ~sum(is.na(.)),
        n_total    = ~length(.),
        pct_na     = ~mean(is.na(.)) * 100
      ),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  ) %>% 
  select(country_wave, ends_with("pct_na")) %>% 
  mutate_if(is.numeric, ~ round(.,2)) %>% 
  rename_with(~str_remove(., "_pct_na"))

num_cols <- setdiff(names(tab), "country_wave")

tab_fmt <- tab %>%
  mutate(across(
    all_of(num_cols),
    ~ ifelse(
      is.na(.x), "",  # opcional: celdas NA como vacío en la salida
      ifelse(
        .x == 100,
        kableExtra::cell_spec(sprintf("%.2f", .x), "html", color = "white", background = "darkred"),
        sprintf("%.2f", .x)
      )
    )
  ))

kableExtra::kable(tab_fmt,
                  format = "html",
                    booktabs= T, 
                    escape = F, 
                    align = 'c') %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", full_width = FALSE))

```

## N by country and wave

```{r}
#| label: tbl-n-country-wave
#| tbl-cap: "N total by country and survey wave"
#| tbl-cap-location: top
#| results: asis
#| echo: false
#| warning: false

tab_n <- db_proc %>%
  count(country, year, name = "n") %>%
  tidyr::pivot_wider(names_from = year, values_from = n, values_fill = 0) %>%
  select(country, `1999`, `2009`, `2019`) %>%
  mutate(Total = `1999` + `2009` + `2019`) %>%
  arrange(country) %>%
  janitor::adorn_totals("row")

kableExtra::kable(tab_n,
                  format = "html",
                  booktabs = T,
                  align = 'c',
                  format.args = list(big.mark = ",")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", full_width = FALSE))

```

**N for 86 valid country-waves: 120.403**

## Market justice

```{r}
db_proc <- db_proc %>%
  mutate(market_justice = (just_health + just_educ) / 2)
```

```{r}
#| label: tbl-market-justice-country-wave
#| tbl-cap: "Market justice index (mean) by country and survey wave"
#| tbl-cap-location: top
#| results: asis
#| echo: false
#| warning: false

tab_mj <- db_proc %>%
  group_by(country, year) %>%
  summarise(market_justice = mean(market_justice, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = year, values_from = market_justice) %>%
  select(country, `1999`, `2009`, `2019`) %>%
  mutate_if(is.numeric, ~ round(., 2)) %>%
  arrange(country)

num_cols_mj <- setdiff(names(tab_mj), "country")

tab_mj_fmt <- tab_mj %>%
  mutate(across(
    all_of(num_cols_mj),
    ~ ifelse(
      is.na(.x), "",
      ifelse(
        .x > 2.50,
        kableExtra::cell_spec(sprintf("%.2f", .x), "html", color = "white", background = "darkred"),
        sprintf("%.2f", .x)
      )
    )
  ))

kableExtra::kable(tab_mj_fmt,
                  format = "html",
                  booktabs = T,
                  escape = F,
                  align = 'c') %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", full_width = FALSE))

```

## Class

```{r}
#| label: fig-egp-wave
#| fig-cap: "Class structure (EGP-5) by ISSP wave"
#| fig-width: 11
#| fig-height: 4.5
#| echo: false
#| warning: false

egp_labels <- c(
  "I-III      White-collar workers" = "White-collar (I-III)",
  "IV.a+IV.b  Petty bourgeoisie"    = "Petty bourgeoisie (IV.a-b)",
  "IV.c+VII.b Farm workers"         = "Farm workers (IV.c+VII.b)",
  "V+VI       Skilled workers"      = "Skilled workers (V-VI)",
  "VII.a      Non-skilled workers"  = "Non-skilled workers (VII.a)"
)

egp_colors <- c(
  "White-collar (I-III)"        = "#2a78d6",
  "Petty bourgeoisie (IV.a-b)"  = "#eb6834",
  "Farm workers (IV.c+VII.b)"   = "#1baf7a",
  "Skilled workers (V-VI)"      = "#eda100",
  "Non-skilled workers (VII.a)" = "#e87ba4"
)

tab_egp <- db_proc %>%
  filter(!is.na(egp5)) %>%
  mutate(egp5_lab = factor(egp_labels[as.character(egp5)], levels = egp_labels)) %>%
  count(year, egp5_lab, name = "n") %>%
  group_by(year) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

ggplot(tab_egp, aes(x = egp5_lab, y = pct, fill = egp5_lab)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.0f%%", pct)), hjust = -0.2, size = 3, color = "#0b0b0b") +
  coord_flip(clip = "off") +
  facet_wrap(~ year, nrow = 1) +
  scale_fill_manual(values = egp_colors, guide = "none") +
  scale_y_continuous(limits = c(0, max(tab_egp$pct) * 1.2), expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "% within wave", title = "Class structure (EGP-5) by ISSP wave") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "#e1e0d9"),
    axis.line.x = element_line(color = "#c3c2b7"),
    axis.text = element_text(color = "#52514e"),
    strip.text = element_text(face = "bold", color = "#0b0b0b"),
    strip.background = element_blank(),
    plot.title = element_text(color = "#0b0b0b", face = "bold"),
    plot.background = element_rect(fill = "#fcfcfb", color = NA),
    panel.background = element_rect(fill = "#fcfcfb", color = NA)
  )

```

## Social subjective status

```{r}
#| label: fig-ess-wave
#| fig-cap: "Subjective social status (ESS) by ISSP wave"
#| fig-width: 8.5
#| fig-height: 5
#| echo: false
#| warning: false

ess_colors <- c("1999" = "#2a78d6", "2009" = "#eb6834", "2019" = "#1baf7a")

tab_ess <- db_proc %>%
  filter(!is.na(ess), ess >= 1, ess <= 10) %>%
  mutate(wave = factor(year)) %>%
  count(wave, ess, name = "n") %>%
  group_by(wave) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

label_df <- tab_ess %>% group_by(wave) %>% filter(ess == max(ess)) %>% ungroup()

ggplot(tab_ess, aes(x = ess, y = pct, color = wave, group = wave)) +
  geom_line(linewidth = 0.75) +
  geom_point(size = 2, fill = "#fcfcfb") +
  ggrepel::geom_text_repel(data = label_df, aes(label = wave), hjust = 0, nudge_x = 0.4,
                            direction = "y", segment.color = NA, size = 3.5, fontface = "bold",
                            show.legend = FALSE) +
  scale_color_manual(values = ess_colors, guide = "none") +
  scale_x_continuous(breaks = 1:10, limits = c(1, 11.3), expand = expansion(mult = c(0.02, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
  labs(x = "Subjective social status (1 = bottom, 10 = top)", y = "% within wave",
       title = "Subjective social status (ESS) by ISSP wave") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#e1e0d9"),
    axis.line = element_line(color = "#c3c2b7"),
    axis.text = element_text(color = "#52514e"),
    plot.title = element_text(color = "#0b0b0b", face = "bold"),
    plot.background = element_rect(fill = "#fcfcfb", color = NA),
    panel.background = element_rect(fill = "#fcfcfb", color = NA)
  )
```

### Meritocratic items

#### Effort

```{r}
#| label: fig-merit-effort-wave
#| fig-cap: "Mean merit_effort by ISSP wave"
#| fig-width: 6.5
#| fig-height: 5.3
#| echo: false
#| warning: false

tab_me <- db_proc %>%
  filter(!is.na(merit_effort)) %>%
  mutate(wave = factor(year)) %>%
  group_by(wave) %>%
  summarise(mean_effort = mean(merit_effort), .groups = "drop")

ggplot(tab_me, aes(x = wave, y = mean_effort)) +
  geom_col(fill = "#2a78d6", width = 0.55) +
  geom_text(aes(label = sprintf("%.2f", mean_effort)), vjust = -0.6, size = 4, color = "#0b0b0b", fontface = "bold") +
  scale_y_continuous(limits = c(0, max(tab_me$mean_effort) * 1.18), expand = expansion(mult = c(0, 0))) +
  labs(x = NULL, y = "Mean merit_effort (1-5)",
       title = "Belief in the importance/reward of effort, by ISSP wave",
       caption = "Note: 1999 measures a different item (\"people get rewarded for their effort\", agree-disagree)\nthan 2009/2019 (\"how important is hard work for getting ahead\"). Not directly comparable.") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "#e1e0d9"),
    axis.line.x = element_line(color = "#c3c2b7"),
    axis.text = element_text(color = "#52514e"),
    plot.title = element_text(color = "#0b0b0b", face = "bold"),
    plot.caption = element_text(color = "#898781", size = 8, hjust = 0, face = "italic"),
    plot.background = element_rect(fill = "#fcfcfb", color = NA),
    panel.background = element_rect(fill = "#fcfcfb", color = NA)
  )

```
