---
title: "Data preparation"
subtitle: "Market justice preferences with ISSP (1987-2019)"
author: "Tomás Urzúa"
date: today
lang: en
fontsize: 12pt
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    toc-expand: 2
    toc-title: Contents
    number-sections: true
    number-depth: 3
    theme:
      - cosmo
      - edumer_html.scss
    code-link: true
    code-fold: true
    title-block-banner: true
  pdf:
    number-sections: true
    number-depth: 3
editor_options: 
  chunk_output_type: console
---

# Presentation

Preparation for 1987, 1992, 1999, 2009 and 2019 database from [Social Inequality Module of International Social Survey Programme (ISSP)](https://www.gesis.org/en/issp/data-and-documentation/social-inequality).

```{r}
#| label: set
#| echo: false
#| message: false
#| warning: false

library(knitr)
knitr::opts_chunk$set(echo = TRUE, include = TRUE, warning = FALSE, message = FALSE)

table_format <- if(is_html_output()) {
  "html"
} else if(is_latex_output()) {
  "latex"
}
table_format2 <- if(is_html_output()) {
  T
} else if(is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")
```


## Libraries

```{r}
#| label: packages
#| include: true

if (! require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse,
               car,
               sjmisc,
               here,
               sjlabelled,
               SciViews,
               naniar,
               haven,
               countrycode,
               ggrepel)


library(occupar)
library(DIGCLASS)
options(scipen=999)
rm(list = ls())
```

## Data

```{r}
#| label: data
#| echo: true
#| output: false

issp87 <- read_dta(here("input/data/original/issp_1987.dta"))
issp92 <- read_dta(here("input/data/original/issp_1992.dta"))
issp99 <- read_dta(here("input/data/original/issp_1999.dta"))
issp09 <- read_dta(here("input/data/original/issp_2009.dta"))
issp19 <- read_dta(here("input/data/original/issp_2019.dta"))
issp_cumu <- read_dta(here("input/data/original/issp_cumulative.dta"))
```

# ISSP 1987

## Select

```{r}

db_proc87 <- issp87 %>%
  select(id = v2,
         country = v3,
         sex = v83,
         age = v82,
         income = v94,
         ess = v65,
         nmerit_wealthy_fam = v4,
         nmerit_educated_parents = v5,
         merit_ambition = v7,
         merit_effort = v9,
         nmerit_contacts = v10,
         nmerit_political_connection = v11,
         nmerit_race = v12,
         nmerit_gender = v15,
         isco68 = v75,
         selfemp = v77,
         wrksup = v78,
         weight = v107) %>%
  janitor::clean_names()

```


##  Recode and transform

### Year

```{r}
db_proc87$year <- 1987
db_proc87$year <- as.numeric(db_proc87$year)
```

### Country

```{r}
frq(db_proc87$country)
db_proc87 <- db_proc87 %>%
  mutate(country = case_when(
    country == 1  ~ "Australia",
    country == 2  ~ "Germany",
    country == 3  ~ "United Kingdom",
    country == 4  ~ "United States",
    country == 5  ~ "Austria",
    country == 6  ~ "Hungary",
    country == 7  ~ "Netherlands",
    country == 8  ~ "Italy",
    country == 11 ~ "Switzerland",
    country == 12 ~ "Poland",
    TRUE ~ NA_character_
  ))
db_proc87$iso_country <- countrycode::countrycode(db_proc87$country, origin = "country.name", destination = "iso3c")
frq(db_proc87$country); frq(db_proc87$iso_country)

```

### Sex
```{r}
frq(db_proc87$sex)
db_proc87$sex <- set_na(db_proc87$sex, na = c(9), drop.levels = F)
db_proc87 <- db_proc87 %>%
mutate(sex = if_else(sex == 1, 'Male', 'Female'),
       sex = factor(sex, levels = c("Male", "Female")))
frq(db_proc87$sex)
```

### Age
```{r}
summary(db_proc87$age)
db_proc87$age <- set_na(db_proc87$age, na = c(99), drop.levels = F)
db_proc87$age <- as.numeric(db_proc87$age)
```

### Education

```{r}
degree87 <- issp_cumu %>%
  filter(year == 1987) %>%
  select(id = caseid, degree = DEGREE)

db_proc87 <- db_proc87 %>%
  left_join(degree87, by = "id")

frq(db_proc87$degree)
db_proc87$degree <- set_na(db_proc87$degree, na = c(-9), drop.levels = F)
db_proc87 <- db_proc87 %>%
  mutate(educ = case_when(degree == 0 ~ "Incomplete primary or less",
                            degree == 1 ~ "Complete primary",
                            degree == 2 ~ "Incomplete secondary",
                            degree == 3 ~ "Complete secondary",
                            degree == 4 ~ "Incomplete university",
                            degree == 5 ~ "Complete university",
                            TRUE ~ NA_character_),
           educ = factor(educ,
                           levels = c("Incomplete primary or less",
                           "Complete primary",
                           "Incomplete secondary",
                           "Complete secondary",
                           "Incomplete university",
                           "Complete university")))

db_proc87 <- db_proc87 %>%
  mutate(educ_dic = case_when(degree %in% c(0:4) ~ 'Less than universitary',
                              degree == 5 ~ 'Universitary',
                              TRUE ~ NA_character_),
         educ_dic = factor(educ_dic, levels = c("Universitary", "Less than universitary")))
frq(db_proc87$educ)
frq(db_proc87$educ_dic)
```

### Income
```{r}

# ntile function without NAs
ntile_zero_na <- function(x, n = 10) {
  out   <- rep(NA_real_, length(x))
  is_na <- is.na(x)
  is0   <- !is_na & x == 0
  pos   <- !is_na & x > 0
  out[is0] <- 0
  if (any(pos)) out[pos] <- ntile(x[pos], n)
  out
}

summary(as.numeric(db_proc87$income))
db_proc87 <- db_proc87 %>%
  mutate(
    income = as.numeric(income),
    income = if_else(country == "Poland" & income == 0, NA_real_, income), # 0 = self-employed (not zero income) in Poland
    income = car::recode(income, "999990:999999 = NA"),
    income_decil = ntile_zero_na(income, 10)
  )

db_proc87 %>%
  group_by(country) %>%
  summarise(n = n(), n_valid = sum(!is.na(income_decil)), .groups = "drop")
frq(db_proc87$income_decil)

```

### Subjective social status
```{r}
frq(db_proc87$ess)
db_proc87$ess <- set_na(db_proc87$ess, na = c(97,98,99), drop.levels = F)
db_proc87$ess <- as.numeric(db_proc87$ess)
frq(db_proc87$ess)
```

### Meritocratic variables
```{r}
frq(db_proc87$merit_effort)
frq(db_proc87$merit_ambition)

db_proc87 <- db_proc87 %>%
  mutate(
    across(
      .cols = c(merit_effort, merit_ambition),
      .fns = ~ set_na(., na = c(0,7,8,9), drop.levels = F)
    )
  ) # ok

db_proc87$merit_effort <- sjmisc::rec(db_proc87$merit_effort,rec = "rev")
db_proc87$merit_ambition <- sjmisc::rec(db_proc87$merit_ambition,rec = "rev")

frq(db_proc87$merit_effort)
frq(db_proc87$merit_ambition)
```

### Non-meritocratic variables
```{r}
frq(db_proc87$nmerit_wealthy_fam)
frq(db_proc87$nmerit_contacts)
frq(db_proc87$nmerit_educated_parents)
frq(db_proc87$nmerit_political_connection)
frq(db_proc87$nmerit_race)
frq(db_proc87$nmerit_gender)

db_proc87 <- db_proc87 %>%
  mutate(
    across(
      .cols = c(nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents,
                nmerit_political_connection, nmerit_race, nmerit_gender),
      .fns = ~ set_na(., na = c(0,7,8,9), drop.levels = F)
    )
  ) # ok

db_proc87$nmerit_wealthy_fam <- sjmisc::rec(db_proc87$nmerit_wealthy_fam,rec = "rev")
db_proc87$nmerit_contacts <- sjmisc::rec(db_proc87$nmerit_contacts,rec = "rev")
db_proc87$nmerit_educated_parents <- sjmisc::rec(db_proc87$nmerit_educated_parents,rec = "rev")
db_proc87$nmerit_political_connection <- sjmisc::rec(db_proc87$nmerit_political_connection,rec = "rev")
db_proc87$nmerit_race <- sjmisc::rec(db_proc87$nmerit_race,rec = "rev")
db_proc87$nmerit_gender <- sjmisc::rec(db_proc87$nmerit_gender,rec = "rev")

frq(db_proc87$nmerit_wealthy_fam)
frq(db_proc87$nmerit_contacts)
frq(db_proc87$nmerit_educated_parents)
frq(db_proc87$nmerit_political_connection)
frq(db_proc87$nmerit_race)
frq(db_proc87$nmerit_gender)

```

### Market justice healthcare
```{r}
# No disponible en ISSP 1987. El ítem "the rich can buy better health care
# than the poor" fue introducido en ISSP 1999 (Social Inequality III).
db_proc87$just_health <- NA_real_
```

### Market justice in education
```{r}
# No disponible en ISSP 1987, misma razón que just_health.
db_proc87$just_educ <- NA_real_
```

### Perceived inequality
```{r}
# No disponible en ISSP 1987. El ítem del diagrama de pirámide social se
# incorporó recién en ISSP 1992.
db_proc87$perc_inequality <- NA_real_
```

### Class

```{r}
db_proc87 <- db_proc87 %>%
  mutate(
    # El codebook solo confirma ISCO68 de 3 dígitos para Alemania, Austria y
    # Suiza (Nota 1); el resto de países usan escalas propias no ISCO.
    isco68 = case_when(country %in% c("Germany", "Austria", "Switzerland") ~ as.numeric(isco68),
                        TRUE ~ NA_real_),
    isco68 = set_na(isco68, na = c(0,999997,999998,999999), drop.levels = F),
    isco88_4 = isco68to88(isco68),
    selfemp = set_na(selfemp, na = c(0,9), drop.levels = F),
    selfemp_bin = if_else(selfemp %in% c(1,2,3), 1, 0), # incl. "self-employed with employees" y (Italia) "self-employed at someone's account"
    nemploy_rec = NA_real_, # no existe variable de número de empleados en 1987; isco88toEGP() asume 0
    egp5 = isco88toEGP(isco88_4, n.employees = nemploy_rec,
                        self.employed = selfemp_bin, n.classes = 5)
  )

frq(db_proc87$egp5)
```

## Save

```{r}
db_proc87 <- db_proc87 %>%
  select(year, country, iso_country, sex, age, educ, educ_dic, income_decil, ess, just_health, just_educ, perc_inequality, egp5, merit_effort, merit_ambition, nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents, nmerit_political_connection, nmerit_race, nmerit_gender, weight)

save(db_proc87, file = here("input/data/proc/db_proc87.RData"))
```

# ISSP 1992

## Select

```{r}

db_proc92 <- issp92 %>%
  select(id = v2,
         country = v3,
         sex = v99,
         age = v100,
         aus117:rp117,
         ess = v73,
         nmerit_wealthy_fam = v4,
         nmerit_educated_parents = v5,
         merit_ambition = v7,
         merit_effort = v9,
         nmerit_contacts = v10,
         nmerit_political_connection = v11,
         nmerit_race = v12,
         nmerit_gender = v15,
         perc_inequality = v90,
         isco68 = v106,
         aus106, pl106, slo106, rus106, nz106,
         selfemp = v110,
         nemploy = v111,
         aus111, d111, i111, n111, rp111,
         wrksup = v107,
         nwrksup = v108,
         weight = v176) %>%
  janitor::clean_names()

```

##  Recode and transform

### Year

```{r}
db_proc92$year <- 1992
db_proc92$year <- as.numeric(db_proc92$year)
```

### Country

```{r}
frq(db_proc92$country)
db_proc92 <- db_proc92 %>%
  mutate(country = case_when(
    country == 1  ~ "Australia",
    country == 2  ~ "Germany",
    country == 3  ~ "Germany",
    country == 4  ~ "United Kingdom",
    country == 5  ~ "United States",
    country == 6  ~ "Austria",
    country == 7  ~ "Hungary",
    country == 8  ~ "Italy",
    country == 9  ~ "Norway",
    country == 10 ~ "Sweden",
    country == 11 ~ "Czechia",
    country == 12 ~ "Slovenia",
    country == 13 ~ "Poland",
    country == 14 ~ "Bulgaria",
    country == 15 ~ "Russia",
    country == 16 ~ "New Zealand",
    country == 17 ~ "Canada",
    country == 18 ~ "Philippines",
    TRUE ~ NA_character_
  ))
db_proc92$iso_country <- countrycode::countrycode(db_proc92$country, origin = "country.name", destination = "iso3c")
frq(db_proc92$country); frq(db_proc92$iso_country)

```

### Sex
```{r}
frq(db_proc92$sex)
db_proc92$sex <- set_na(db_proc92$sex, na = c(9), drop.levels = F)
db_proc92 <- db_proc92 %>%
mutate(sex = if_else(sex == 1, 'Male', 'Female'),
       sex = factor(sex, levels = c("Male", "Female")))
frq(db_proc92$sex)
```

### Age
```{r}
summary(db_proc92$age)
db_proc92$age <- set_na(db_proc92$age, na = c(99), drop.levels = F)
db_proc92$age <- as.numeric(db_proc92$age)
```

### Education

```{r}
degree92 <- issp_cumu %>%
  filter(year == 1992) %>%
  select(id = caseid, degree = DEGREE)

db_proc92 <- db_proc92 %>%
  left_join(degree92, by = "id")

frq(db_proc92$degree)
db_proc92$degree <- set_na(db_proc92$degree, na = c(-9), drop.levels = F)
db_proc92 <- db_proc92 %>%
  mutate(educ = case_when(degree == 0 ~ "Incomplete primary or less",
                            degree == 1 ~ "Complete primary",
                            degree == 2 ~ "Incomplete secondary",
                            degree == 3 ~ "Complete secondary",
                            degree == 4 ~ "Incomplete university",
                            degree == 5 ~ "Complete university",
                            TRUE ~ NA_character_),
           educ = factor(educ,
                           levels = c("Incomplete primary or less",
                           "Complete primary",
                           "Incomplete secondary",
                           "Complete secondary",
                           "Incomplete university",
                           "Complete university")))

db_proc92 <- db_proc92 %>%
  mutate(educ_dic = case_when(degree %in% c(0:4) ~ 'Less than universitary',
                              degree == 5 ~ 'Universitary',
                              TRUE ~ NA_character_),
         educ_dic = factor(educ_dic, levels = c("Universitary", "Less than universitary")))
frq(db_proc92$educ)
frq(db_proc92$educ_dic)
```



### Income
```{r}

# ntile function without NAs
ntile_zero_na <- function(x, n = 10) {
  out   <- rep(NA_real_, length(x))
  is_na <- is.na(x)
  is0   <- !is_na & x == 0
  pos   <- !is_na & x > 0
  out[is0] <- 0
  if (any(pos)) out[pos] <- ntile(x[pos], n)
  out
}

db_proc92 <- db_proc92 %>%
  mutate_at(vars(aus117:rp117), ~ as.numeric(.)) %>%
  mutate_at(vars(aus117:rp117), funs(car::recode(., "999997 = NA; 999998 = NA; 999999 = NA"))) %>%
  mutate(
    income_personal = do.call(coalesce, across(aus117:rp117)),
    income_decil    = ntile_zero_na(income_personal, 10)
  )

db_proc92 %>%
  mutate(n_non_na = rowSums(!is.na(pick(aus117:rp117)))) %>%
  count(n_non_na)
frq(db_proc92$income_decil)

```

### Subjective social status
```{r}
frq(db_proc92$ess)
db_proc92$ess <- set_na(db_proc92$ess, na = c(97,98,99), drop.levels = F)
db_proc92$ess <- as.numeric(db_proc92$ess)
frq(db_proc92$ess)
```

### Meritocratic variables
```{r}
frq(db_proc92$merit_effort)
frq(db_proc92$merit_ambition)

db_proc92 <- db_proc92 %>%
  mutate(
    across(
      .cols = c(merit_effort, merit_ambition),
      .fns = ~ set_na(., na = c(8,9), drop.levels = F)
    )
  ) # ok

db_proc92$merit_effort <- sjmisc::rec(db_proc92$merit_effort,rec = "rev")
db_proc92$merit_ambition <- sjmisc::rec(db_proc92$merit_ambition,rec = "rev")

frq(db_proc92$merit_effort)
frq(db_proc92$merit_ambition)
```

### Non-meritocratic variables
```{r}
frq(db_proc92$nmerit_wealthy_fam)
frq(db_proc92$nmerit_contacts)
frq(db_proc92$nmerit_educated_parents)
frq(db_proc92$nmerit_political_connection)
frq(db_proc92$nmerit_race)
frq(db_proc92$nmerit_gender)

db_proc92 <- db_proc92 %>%
  mutate(
    across(
      .cols = c(nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents,
                nmerit_political_connection, nmerit_race, nmerit_gender),
      .fns = ~ set_na(., na = c(8,9), drop.levels = F)
    )
  ) # ok

db_proc92$nmerit_wealthy_fam <- sjmisc::rec(db_proc92$nmerit_wealthy_fam,rec = "rev")
db_proc92$nmerit_contacts <- sjmisc::rec(db_proc92$nmerit_contacts,rec = "rev")
db_proc92$nmerit_educated_parents <- sjmisc::rec(db_proc92$nmerit_educated_parents,rec = "rev")
db_proc92$nmerit_political_connection <- sjmisc::rec(db_proc92$nmerit_political_connection,rec = "rev")
db_proc92$nmerit_race <- sjmisc::rec(db_proc92$nmerit_race,rec = "rev")
db_proc92$nmerit_gender <- sjmisc::rec(db_proc92$nmerit_gender,rec = "rev")

frq(db_proc92$nmerit_wealthy_fam)
frq(db_proc92$nmerit_contacts)
frq(db_proc92$nmerit_educated_parents)
frq(db_proc92$nmerit_political_connection)
frq(db_proc92$nmerit_race)
frq(db_proc92$nmerit_gender)

```

### Market justice healthcare
```{r}
# No disponible en ISSP 1992 (Social Inequality II). El ítem "the rich can buy
# better health care than the poor" fue introducido en ISSP 1999 (Social
# Inequality III).
db_proc92$just_health <- NA_real_
```

### Market justice in education
```{r}
# No disponible en ISSP 1992, misma razón que just_health.
db_proc92$just_educ <- NA_real_
```

### Perceived inequality
```{r}
frq(db_proc92$perc_inequality)
db_proc92$perc_inequality <- set_na(db_proc92$perc_inequality, na = c(0,8,9), drop.levels = F)
db_proc92$perc_inequality <- sjmisc::rec(db_proc92$perc_inequality,rec = "rev")
frq(db_proc92$perc_inequality)
```

### Class

```{r}
db_proc92 <- db_proc92 %>%
  mutate(
    isco68 = set_na(isco68, na = c(0,9997,9998,9999), drop.levels = F),
    isco88_conv = isco68to88(isco68),
    isco88_cty = coalesce(
      na_if(as.numeric(aus106), 0),
      na_if(as.numeric(pl106), 0),
      na_if(as.numeric(slo106), 0),
      na_if(as.numeric(rus106), 0),
      na_if(as.numeric(nz106), 0)
    ),
    isco88_4 = coalesce(isco88_conv, isco88_cty),
    selfemp = set_na(selfemp, na = c(0,9), drop.levels = F),
    selfemp_bin = if_else(selfemp == 1, 1, 0),
    nemploy = set_na(nemploy, na = c(99), drop.levels = F),
    nemploy_cty = coalesce(
      na_if(as.numeric(aus111), 0),
      na_if(as.numeric(d111), 0),
      na_if(as.numeric(i111), 0),
      na_if(as.numeric(n111), 0),
      na_if(as.numeric(rp111), 0)
    ),
    nemploy_rec = coalesce(as.numeric(nemploy), nemploy_cty),
    egp5 = isco88toEGP(isco88_4, n.employees = nemploy_rec,
                        self.employed = selfemp_bin, n.classes = 5)
  )

frq(db_proc92$egp5)
```

## Save

```{r}
db_proc92 <- db_proc92 %>%
  select(year, country, iso_country, sex, age, educ, educ_dic, income_decil, ess, just_health, just_educ, perc_inequality, egp5, merit_effort, merit_ambition, nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents, nmerit_political_connection, nmerit_race, nmerit_gender, weight)

save(db_proc92, file = here("input/data/proc/db_proc92.RData"))
```



# ISSP 1999

## Select

```{r}

db_proc99 <- issp99 %>% 
  select(year = v1, 
         v3,
         sex,
         age,
         educyrs,
         degree,
         rincome,
         rincomer,
         ess = v46,
         nmerit_wealthy_fam = v4,
         nmerit_contacts = v5,
         merit_effort = v6,
         merit_talent = v7,
         just_health = v39,
         just_educ = v40,
         perc_inequality = v57,
         isco88_4,
         selfemp,
         nemploy,
         wrksup,
         nwrksup,
         weight) %>% 
  janitor::clean_names()

```


##  Recode and transform

### Year

```{r}
frq(db_proc99$year)
db_proc99$year <- as.numeric(db_proc99$year)
frq(db_proc99$year)
```

### Country

```{r}
frq(db_proc99$v3)
db_proc99 <- db_proc99 %>% 
  mutate(country = case_when(
    v3 == 1  ~ "Australia",
    v3 == 2 | v3 == 3  ~ "Germany",
    v3 == 4  ~ "United Kingdom",
    v3 == 5  ~ "Ireland",
    v3 == 6  ~ "United States",
    v3 == 7  ~ "Austria",
    v3 == 8  ~ "Hungary",
    v3 == 10 ~ "Ireland",      
    v3 == 12 ~ "Norway",
    v3 == 13 ~ "Sweden",
    v3 == 14 ~ "Czechia",  
    v3 == 15 ~ "Slovenia",
    v3 == 16 ~ "Poland",
    v3 == 17 ~ "Bulgaria",
    v3 == 18 ~ "Russia",
    v3 == 19 ~ "New Zealand",
    v3 == 20 ~ "Canada",
    v3 == 21 ~ "Philippines",
    v3 == 22 ~ "Israel",
    v3 == 24 ~ "Japan",
    v3 == 25 ~ "Spain",
    v3 == 26 ~ "Latvia",
    v3 == 27 ~ "France",
    v3 == 28 ~ "Cyprus",
    v3 == 29 ~ "Portugal",
    v3 == 30 ~ "Chile",
    v3 == 33 ~ "Slovakia",
    TRUE ~ NA_character_
  ))
db_proc99$iso_country <- countrycode::countrycode(db_proc99$country, origin = "country.name", destination = "iso3c")
frq(db_proc99$country); frq(db_proc99$iso_country)

```

### Sex
```{r}
frq(db_proc99$sex)
db_proc99$sex <- set_na(db_proc99$sex, na = c(9), drop.levels = F)
db_proc99 <- db_proc99 %>% 
mutate(sex = if_else(sex == 1, 'Male', 'Female'),
       sex = factor(sex, levels = c("Male", "Female")))
frq(db_proc99$sex)
```


### Age
```{r}
summary(db_proc99$age)
db_proc99$age <- set_na(db_proc99$age, na = c(98,99), drop.levels = F)
db_proc99$age <- as.numeric(db_proc99$age)

```

### Education
```{r}
# Eucation
frq(db_proc99$educyrs) # We don't know if this is equal in every country; we use degree
frq(db_proc99$degree)
db_proc99$degree <- set_na(db_proc99$degree, na = c(8,9), drop.levels = F)
db_proc99 <- db_proc99 %>% 
  mutate(educ = case_when(degree %in% c(1,2) ~ "Incomplete primary or less",
                            degree == 3 ~ "Complete primary",
                            degree == 4 ~ "Incomplete secondary",
                            degree == 5 ~ "Complete secondary",
                            degree == 6 ~ "Incomplete university",
                            degree == 7 ~ "Complete university",
                            TRUE ~ NA_character_),
           educ = factor(educ, 
                           levels = c("Incomplete primary or less",
                           "Complete primary",
                           "Incomplete secondary",
                           "Complete secondary",
                           "Incomplete university",
                           "Complete university")))

db_proc99 <- db_proc99 %>% 
  mutate(educ_dic = case_when(degree %in% c(1:6) ~ 'Less than universitary',
                              degree == 7 ~ 'Universitary',
                              TRUE ~ NA_character_),
         educ_dic = factor(educ_dic, levels = c("Universitary", "Less than universitary")))
frq(db_proc99$educ)
frq(db_proc99$educ_dic)
```


### Income
```{r}

# Income continuous
summary(db_proc99$rincome)
db_proc99 <- db_proc99 %>%
  mutate(
    rincome = haven::zap_labels(rincome),
    rincome = sjmisc::set_na(rincome, na = c(999997,999998,999999))
  )

# ntile function without NAs
ntile_na <- function(x,n)
{
  notna <- !is.na(x)
  out <- rep(NA_real_,length(x))
  out[notna] <- ntile(x[notna],n)
  return(out)
}

db_proc99 %>% 
  mutate(decil = ntile_na(rincome, 10)) %>% 
  frq(decil)

# Income decil
frq(db_proc99$rincomer) 
db_proc99$rincomer <- set_na(db_proc99$rincomer, na = c(97,98,99), drop.levels = F) # there is a decil 0; those no income no labor force
db_proc99$income_decil <- as.numeric(db_proc99$rincomer)
frq(db_proc99$income_decil) 
```

### Subjective social status
```{r}
frq(db_proc99$ess)
db_proc99$ess <- set_na(db_proc99$ess, na = c(97,98,99), drop.levels = F)
db_proc99$ess <- as.numeric(db_proc99$ess)
frq(db_proc99$ess)
```

### Meritocratic variables
```{r}
frq(db_proc99$merit_effort)
frq(db_proc99$merit_talent)

db_proc99 <- db_proc99 %>% 
  mutate(
    across(
      .cols = c(merit_effort, merit_talent),
      .fns = ~ set_na(., na = c(8,9), drop.levels = F)
    )
  ) # ok

db_proc99$merit_effort <- sjmisc::rec(db_proc99$merit_effort,rec = "rev")
db_proc99$merit_talent <- sjmisc::rec(db_proc99$merit_talent,rec = "rev")

frq(db_proc99$merit_effort)
frq(db_proc99$merit_talent)
```

### Non-meritocratic variables
```{r}
frq(db_proc99$nmerit_wealthy_fam)
frq(db_proc99$nmerit_contacts)

db_proc99 <- db_proc99 %>% 
  mutate(
    across(
      .cols = c(nmerit_wealthy_fam, nmerit_contacts),
      .fns = ~ set_na(., na = c(8,9), drop.levels = F)
    )
  ) # ok

db_proc99$nmerit_wealthy_fam <- sjmisc::rec(db_proc99$nmerit_wealthy_fam,rec = "rev")
db_proc99$nmerit_contacts <- sjmisc::rec(db_proc99$nmerit_contacts,rec = "rev")

frq(db_proc99$nmerit_wealthy_fam)
frq(db_proc99$nmerit_contacts)
```

### Market justice healthcare
```{r}
frq(db_proc99$just_health)
db_proc99$just_health <- set_na(db_proc99$just_health, na = c(8,9), drop.levels = F) 
db_proc99$just_health <- sjmisc::rec(db_proc99$just_health,rec = "rev")
frq(db_proc99$just_health)
?rec
```

### Market justice in education
```{r}
frq(db_proc99$just_educ)
db_proc99$just_educ <- set_na(db_proc99$just_educ, na = c(8,9), drop.levels = F) 
db_proc99$just_educ <- sjmisc::rec(db_proc99$just_educ,rec = "rev")
frq(db_proc99$just_educ)

```

### Perceived inequality
```{r}
frq(db_proc99$perc_inequality)
db_proc99$perc_inequality <- set_na(db_proc99$perc_inequality, na = c(8,9), drop.levels = F)
db_proc99$perc_inequality <- sjmisc::rec(db_proc99$perc_inequality,rec = "rev")
frq(db_proc99$perc_inequality)
```

### Class

```{r}
db_proc99 <- db_proc99 %>%
  mutate(
    selfemp_bin = if_else(selfemp == 1, 1, 0),
    nemploy_rec = case_when(
      nemploy == 9995 ~ 0,
      nemploy %in% c(9997, 9999) ~ NA_real_,
      TRUE ~ nemploy
    ),
    egp5 = isco88toEGP(isco88_4, n.employees = nemploy_rec,
                        self.employed = selfemp_bin, n.classes = 5)
  )

frq(db_proc99$egp5)
```

```{r}
#| label: class-digclass-1999

db_proc99 <- db_proc99 %>%
  mutate(
    isco88_chr = sprintf("%04d", as.numeric(isco88_4)),
    wrksup = set_na(wrksup, na = c(0, 7, 8, 9), drop.levels = F),
    wrksup_bin = if_else(wrksup == 1, 1, 0),
    # Ver nota sobre el bug de indexacion de DIGCLASS::isco88_to_egp() en el
    # chunk "egp schema" (seccion Cumulative data): para independientes se usa
    # 1 en vez de 0 cuando no se conocen empleados, para no caer en la columna
    # equivocada de la tabla de equivalencias.
    n_employees_egp = case_when(
      selfemp_bin == 1 & (is.na(nemploy_rec) | nemploy_rec == 0) ~ 1,
      selfemp_bin == 1 ~ nemploy_rec,
      TRUE ~ 0
    ),
    n_employees_esec = if_else(selfemp_bin == 1, coalesce(nemploy_rec, 0), 0),
    egp5_dc_code = DIGCLASS::isco88_to_egp(
      isco88_chr,
      self_employed = selfemp_bin,
      n_employees = n_employees_egp,
      n_classes = 5,
      label = FALSE
    ),
    egp5_digclass = factor(
      egp5_dc_code,
      levels = as.character(1:5),
      labels = c("White-collar (I-III)",
                 "Petty bourgeoisie (IV.a-b)",
                 "Farm workers (IV.c+VII.b)",
                 "Skilled workers (V-VI)",
                 "Non-skilled workers (VII.a)")
    ),
    isco88com = DIGCLASS::isco88_to_isco88com(isco88_chr),
    isco88com_3 = DIGCLASS::isco88_swap(isco88com, from = 4, to = 3),
    esec9_dc = DIGCLASS::isco88com_to_esec(
      isco88com_3,
      is_supervisor = wrksup_bin,
      self_employed = selfemp_bin,
      n_employees = n_employees_esec,
      full_method = TRUE
    ),
    # Colapso oficial ESeC-9 -> ESeC-5, ver chunk "esec schema" (seccion
    # Cumulative data) para la referencia (Harrison & Rose 2006, ESeC User
    # Guide, secciones 3.3-3.4).
    esec5_digclass = case_when(
      esec9_dc %in% c("1", "2") ~ "1",
      esec9_dc %in% c("3", "6") ~ "2",
      esec9_dc %in% c("4", "5") ~ "3",
      esec9_dc == "7"           ~ "4",
      esec9_dc %in% c("8", "9") ~ "5"
    ),
    esec5_digclass = factor(
      esec5_digclass,
      levels = as.character(1:5),
      labels = c("Salariat",
                 "Intermediate employees",
                 "Small employers and self-employed",
                 "Lower services, sales and clerical",
                 "Lower technical and routine")
    )
  )

frq(db_proc99$egp5_digclass)
frq(db_proc99$esec5_digclass)
```

## Save

```{r}
db_proc99 <- db_proc99 %>%
  select(year, country, iso_country, sex, age, educ, educ_dic, income_decil, ess, just_health, just_educ, perc_inequality, egp5, egp5_digclass, esec5_digclass, merit_effort, merit_talent, nmerit_wealthy_fam, nmerit_contacts, weight)

save(db_proc99, file = here("input/data/proc/db_proc99.RData"))
```

# Processing ISSP 2009

## Select

```{r}
db_proc09 <- issp09 %>% 
  janitor::clean_names() %>% 
  select(v5, 
         sex,
         age,
         degree,
         141:181,
         ess = v44,
         nmerit_wealthy_fam = v6, 
         nmerit_educated_parents = v7, 
         merit_ambition = v9, 
         merit_effort = v10,
         nmerit_contacts = v11,
         nmerit_political_connection = v12,
         nmerit_race = v14,
         nmerit_gender = v16,
         just_health = v38,
         just_educ = v39,
         perc_inequality = v54,
         isco88,
         wrktype,
         nemploy,
         wrksup,
         weight
         )

```


##  Recode and transform

### Year

```{r}
db_proc09$year <- 2009
db_proc09$year <- as.numeric(db_proc09$year)
```

### Country

```{r}
frq(db_proc09$v5)
db_proc09$country <- NA
db_proc09$country <- countrycode::countrycode(db_proc09$v5, 
                                              origin = "iso3n", 
                                              destination = "country.name")
db_proc09$iso_country <- countrycode::countrycode(db_proc09$country, 
                                                  origin = "country.name",
                                                  destination = "iso3c")
frq(db_proc09$country); frq(db_proc09$iso_country)

```

### Sex
```{r}
frq(db_proc09$sex)
db_proc09$sex <- set_na(db_proc09$sex, na = c(9), drop.levels = F)
db_proc09 <- db_proc09 %>% 
  mutate(sex = if_else(sex == 1, 'Male', 'Female'),
       sex = factor(sex, levels = c("Male", "Female")))
frq(db_proc09$sex)
```


### Age
```{r}
summary(db_proc09$age)
db_proc09$age <- set_na(db_proc09$age, na = c(99), drop.levels = F)
db_proc09$age <- as.numeric(db_proc09$age)
```

### Education
```{r}
frq(db_proc09$degree)
db_proc09$degree <- set_na(db_proc09$degree, na = c(8,9), drop.levels = F)
db_proc09 <- db_proc09 %>% 
  mutate(educ = case_when(degree == 0 ~ "Incomplete primary or less",
                            degree == 1 ~ "Complete primary",
                            degree == 2 ~ "Incomplete secondary",
                            degree == 3 ~ "Complete secondary",
                            degree == 4 ~ "Incomplete university",
                            degree == 5 ~ "Complete university",
                            TRUE ~ NA_character_),
           educ = factor(educ, 
                           levels = c("Incomplete primary or less",
                           "Complete primary",
                           "Incomplete secondary",
                           "Complete secondary",
                           "Incomplete university",
                           "Complete university")))


db_proc09 <- db_proc09 %>% 
  mutate(educ_dic = case_when(degree %in% c(0:4) ~ 'Less than universitary',
                              degree == 5 ~ 'Universitary',
                              TRUE ~ NA_character_),
         educ_dic = factor(educ_dic, levels = c("Universitary", "Less than universitary")))
frq(db_proc09$educ)
frq(db_proc09$educ_dic)
```


### Income
```{r}
#frq(db_proc09$ar_rinc)
#frq(db_proc09$at_rinc)
#frq(db_proc09$au_rinc)
#frq(db_proc09$be_rinc)
#frq(db_proc09$bg_rinc)
#frq(db_proc09$ch_rinc)
#frq(db_proc09$cl_rinc)
#frq(db_proc09$cn_rinc)
#frq(db_proc09$cy_rinc)
#frq(db_proc09$cz_rinc)
#frq(db_proc09$de_rinc)
#frq(db_proc09$dk_rinc)
#frq(db_proc09$ee_rinc)
#frq(db_proc09$es_rinc)
#frq(db_proc09$fi_rinc)
#frq(db_proc09$fr_rinc)
#frq(db_proc09$gb_rinc)
#frq(db_proc09$hr_rinc)
#frq(db_proc09$hu_rinc)
#frq(db_proc09$il_rinc)
#frq(db_proc09$is_rinc)
#frq(db_proc09$it_rinc)
#frq(db_proc09$jp_rinc)
#frq(db_proc09$kr_rinc)
#frq(db_proc09$lt_rinc)
#frq(db_proc09$lv_rinc)
#frq(db_proc09$no_rinc)
#frq(db_proc09$nz_rinc)
#frq(db_proc09$ph_rinc)
#frq(db_proc09$pl_rinc)
#frq(db_proc09$pt_rinc)
#frq(db_proc09$ru_rinc)
#frq(db_proc09$se_rinc)
#frq(db_proc09$si_rinc)
#frq(db_proc09$sk_rinc)
#frq(db_proc09$tr_rinc)
#frq(db_proc09$tw_rinc)
#frq(db_proc09$ua_rinc)
#frq(db_proc09$us_rinc)
#frq(db_proc09$ve_rinc)
#frq(db_proc09$za_rinc)

# ntile function without NAs
ntile_zero_na <- function(x, n = 10) {
  out   <- rep(NA_real_, length(x))
  is_na <- is.na(x)
  is0   <- !is_na & x == 0
  pos   <- !is_na & x > 0
  out[is0] <- 0
  if (any(pos)) out[pos] <- ntile(x[pos], n)
  out
}

db_proc09 <- db_proc09 %>% 
 mutate_at(vars(6:46), ~ as.numeric(.)) %>% 
  mutate_at(vars(6:46), funs(car::recode(. ,"999990 = NA; 999997 = NA; 999998 = NA; 999999 = NA; 
                                              9999990 = NA; 9999997 = NA; 9999998 = NA; 9999999 = NA;
                                              99999990 = NA; 99999997 = NA; 99999998 = NA; 99999999 = NA"))) %>% 
  mutate(
    income_personal = do.call(coalesce, across(6:46)),
    income_decil    = ntile_zero_na(income_personal, 10)
  )

db_proc09 %>% 
  mutate(n_non_na = rowSums(!is.na(pick(6:46)))) %>% 
  count(n_non_na)
frq(db_proc09$income_decil)

```

### Subjective social status
```{r}
frq(db_proc09$ess)
db_proc09$ess <- set_na(db_proc09$ess, na = c(97,98,99), drop.levels = F)
db_proc09$ess <- as.numeric(db_proc09$ess)
frq(db_proc09$ess)
```

### Meritocratic variables
```{r}
frq(db_proc09$merit_effort)
frq(db_proc09$merit_ambition)

db_proc09 <- db_proc09 %>% 
  mutate(
    across(
      .cols = c(merit_effort, merit_ambition),
      .fns = ~ set_na(., na = c(8,9), drop.levels = F)
    )
  ) # ok

db_proc09$merit_effort <- sjmisc::rec(db_proc09$merit_effort,rec = "rev")
db_proc09$merit_ambition <- sjmisc::rec(db_proc09$merit_ambition,rec = "rev")

frq(db_proc09$merit_effort)
frq(db_proc09$merit_ambition)
```

### Non-meritocratic variables
```{r}
frq(db_proc09$nmerit_wealthy_fam)
frq(db_proc09$nmerit_contacts)
frq(db_proc09$nmerit_educated_parents)
frq(db_proc09$nmerit_political_connection)
frq(db_proc09$nmerit_race)
frq(db_proc09$nmerit_gender)

db_proc09 <- db_proc09 %>% 
  mutate(
    across(
      .cols = c(nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents,
                nmerit_political_connection, nmerit_race, nmerit_gender),
      .fns = ~ set_na(., na = c(8,9), drop.levels = F)
    )
  ) # ok

db_proc09$nmerit_wealthy_fam <- sjmisc::rec(db_proc09$nmerit_wealthy_fam,rec = "rev")
db_proc09$nmerit_contacts <- sjmisc::rec(db_proc09$nmerit_contacts,rec = "rev")
db_proc09$nmerit_educated_parents <- sjmisc::rec(db_proc09$nmerit_educated_parents,rec = "rev")
db_proc09$nmerit_political_connection <- sjmisc::rec(db_proc09$nmerit_political_connection,rec = "rev")
db_proc09$nmerit_race <- sjmisc::rec(db_proc09$nmerit_race,rec = "rev")
db_proc09$nmerit_gender <- sjmisc::rec(db_proc09$nmerit_gender,rec = "rev")

frq(db_proc09$nmerit_wealthy_fam)
frq(db_proc09$nmerit_contacts)
frq(db_proc09$nmerit_educated_parents)
frq(db_proc09$nmerit_political_connection)
frq(db_proc09$nmerit_race)
frq(db_proc09$nmerit_gender)

```

### Market justice healthcare
```{r}
frq(db_proc09$just_health)
db_proc09$just_health <- set_na(db_proc09$just_health, na = c(8,9), drop.levels = F) 
db_proc09$just_health <- sjmisc::rec(db_proc09$just_health,rec = "rev")
frq(db_proc09$just_health)

```

### Market justice in education
```{r}
frq(db_proc09$just_educ)
db_proc09$just_educ <- set_na(db_proc09$just_educ, na = c(8,9), drop.levels = F) 
db_proc09$just_educ <- sjmisc::rec(db_proc09$just_educ,rec = "rev")
frq(db_proc09$just_educ)

```

### Perceived inequality
```{r}
frq(db_proc09$perc_inequality)
db_proc09$perc_inequality <- set_na(db_proc09$perc_inequality, na = c(8,9), drop.levels = F)
db_proc09$perc_inequality <- sjmisc::rec(db_proc09$perc_inequality,rec = "rev")
frq(db_proc09$perc_inequality)
```

### Class

```{r}

db_proc09 <- db_proc09 %>% 
  mutate(
    selfemp_bin = if_else(wrktype == 4, 1, 0),
    nemploy_rec = case_when(
      nemploy == 9995 ~ 0,
      nemploy %in% c(9997, 9998, 9999) ~ NA_real_,
      TRUE ~ nemploy
    ),
    egp5  = isco88toEGP(isco88, n.employees = nemploy_rec,
                         self.employed = selfemp_bin, n.classes = 5)
  )

frq(db_proc09$egp5)
```

```{r}
#| label: class-digclass-2009

db_proc09 <- db_proc09 %>%
  mutate(
    isco88_chr = sprintf("%04d", as.numeric(isco88)),
    wrksup = set_na(wrksup, na = c(0, 7, 8, 9), drop.levels = F),
    wrksup_bin = if_else(wrksup == 1, 1, 0),
    # Ver nota sobre el bug de indexacion de DIGCLASS::isco88_to_egp() en el
    # chunk "egp schema" (seccion Cumulative data): para independientes se usa
    # 1 en vez de 0 cuando no se conocen empleados, para no caer en la columna
    # equivocada de la tabla de equivalencias.
    n_employees_egp = case_when(
      selfemp_bin == 1 & (is.na(nemploy_rec) | nemploy_rec == 0) ~ 1,
      selfemp_bin == 1 ~ nemploy_rec,
      TRUE ~ 0
    ),
    n_employees_esec = if_else(selfemp_bin == 1, coalesce(nemploy_rec, 0), 0),
    egp5_dc_code = DIGCLASS::isco88_to_egp(
      isco88_chr,
      self_employed = selfemp_bin,
      n_employees = n_employees_egp,
      n_classes = 5,
      label = FALSE
    ),
    egp5_digclass = factor(
      egp5_dc_code,
      levels = as.character(1:5),
      labels = c("White-collar (I-III)",
                 "Petty bourgeoisie (IV.a-b)",
                 "Farm workers (IV.c+VII.b)",
                 "Skilled workers (V-VI)",
                 "Non-skilled workers (VII.a)")
    ),
    isco88com = DIGCLASS::isco88_to_isco88com(isco88_chr),
    isco88com_3 = DIGCLASS::isco88_swap(isco88com, from = 4, to = 3),
    esec9_dc = DIGCLASS::isco88com_to_esec(
      isco88com_3,
      is_supervisor = wrksup_bin,
      self_employed = selfemp_bin,
      n_employees = n_employees_esec,
      full_method = TRUE
    ),
    # Colapso oficial ESeC-9 -> ESeC-5, ver chunk "esec schema" (seccion
    # Cumulative data) para la referencia (Harrison & Rose 2006, ESeC User
    # Guide, secciones 3.3-3.4).
    esec5_digclass = case_when(
      esec9_dc %in% c("1", "2") ~ "1",
      esec9_dc %in% c("3", "6") ~ "2",
      esec9_dc %in% c("4", "5") ~ "3",
      esec9_dc == "7"           ~ "4",
      esec9_dc %in% c("8", "9") ~ "5"
    ),
    esec5_digclass = factor(
      esec5_digclass,
      levels = as.character(1:5),
      labels = c("Salariat",
                 "Intermediate employees",
                 "Small employers and self-employed",
                 "Lower services, sales and clerical",
                 "Lower technical and routine")
    )
  )

frq(db_proc09$egp5_digclass)
frq(db_proc09$esec5_digclass)
```

## Save

```{r}
#| collapse: false

db_proc09 <- db_proc09 %>%
  select(year, country, iso_country, sex, age, educ, educ_dic, income_decil, ess, just_health, just_educ, perc_inequality, egp5, egp5_digclass, esec5_digclass, merit_effort, merit_ambition, nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents, nmerit_political_connection, nmerit_race, nmerit_gender, weight)
```


```{r}
save(db_proc09, file = here("input/data/proc/db_proc09.RData"))
```

# Processing ISSP 2019

## Select

```{r}
db_proc19 <- issp19 %>% 
  janitor::clean_names() %>% 
  select(country, 
         sex,
         age,
         degree,
         258:286,
         ess = topbot,
         nmerit_wealthy_fam = v1, 
         nmerit_educated_parents = v2, 
         merit_effort = v4,
         nmerit_contacts = v5,
         nmerit_political_connection = v6,
         nmerit_race = v8,
         nmerit_gender = v10,
         just_health = v30,
         just_educ = v31,
         perc_inequality = v48,
         isco08,
         emprel,
         wrksup,
         nsup,
         weight
         )
```


##  Recode and transform

### Year

```{r}
db_proc19$year <- 2019
db_proc19$year <- as.numeric(db_proc19$year)
```

### Country

```{r}
frq(db_proc19$country)
db_proc19$country <- countrycode::countrycode(db_proc19$country, 
                                              origin = "iso3n", 
                                              destination = "country.name")
db_proc19$iso_country <- countrycode::countrycode(db_proc19$country, 
                                                  origin = "country.name",
                                                  destination = "iso3c")
frq(db_proc19$country); frq(db_proc19$iso_country)

```

### Sex
```{r}
frq(db_proc19$sex)
db_proc19$sex <- set_na(db_proc19$sex, na = c(-9), drop.levels = F)
db_proc19 <- db_proc19 %>% 
  mutate(sex = if_else(sex == 1, 'Male', 'Female'),
       sex = factor(sex, levels = c("Male", "Female")))
frq(db_proc19$sex)
```


### Age
```{r}
summary(db_proc19$age)
db_proc19$age <- set_na(db_proc19$age, na = c(-9), drop.levels = F)
db_proc19$age <- as.numeric(db_proc19$age)
```

### Education
```{r}
frq(db_proc19$degree)
db_proc19$degree <- set_na(db_proc19$degree, na = c(-9), drop.levels = F)
db_proc19 <- db_proc19 %>% 
  mutate(educ = case_when(degree == 0 ~ "Incomplete primary or less",
                            degree == 1 ~ "Complete primary",
                            degree == 2 ~ "Incomplete secondary",
                            degree == 3 ~ "Complete secondary",
                            degree == 4 ~ "Incomplete university",
                            degree %in% c(5,6) ~ "Complete university",
                            TRUE ~ NA_character_),
           educ = factor(educ, 
                           levels = c("Incomplete primary or less",
                           "Complete primary",
                           "Incomplete secondary",
                           "Complete secondary",
                           "Incomplete university",
                           "Complete university")))

db_proc19 <- db_proc19 %>% 
  mutate(educ_dic = case_when(degree %in% c(0:4) ~ 'Less than universitary',
                              degree %in% c(5:6) ~ 'Universitary',
                              TRUE ~ NA_character_),
         educ_dic = factor(educ_dic, levels = c("Universitary", "Less than universitary")))
frq(db_proc19$educ)
frq(db_proc19$educ_dic)
```


### Income
```{r}
#frq(db_proc19$at_rinc)
#frq(db_proc19$au_rinc)
#frq(db_proc19$bg_rinc)
#frq(db_proc19$ch_rinc)
#frq(db_proc19$cl_rinc)
#frq(db_proc19$cz_rinc)
#frq(db_proc19$de_rinc)
#frq(db_proc19$dk_rinc)
#frq(db_proc19$fi_rinc)
#frq(db_proc19$fr_rinc)
#frq(db_proc19$gb_rinc)
#frq(db_proc19$hr_rinc)
#frq(db_proc19$il_rinc)
#frq(db_proc19$is_rinc)
#frq(db_proc19$it_rinc)
#frq(db_proc19$jp_rinc)
#frq(db_proc19$lt_rinc)
#frq(db_proc19$no_rinc)
#frq(db_proc19$nz_rinc)
#frq(db_proc19$ph_rinc)
#frq(db_proc19$ru_rinc)
#frq(db_proc19$se_rinc)
#frq(db_proc19$si_rinc)
#frq(db_proc19$sr_rinc)
#frq(db_proc19$th_rinc)
#frq(db_proc19$tw_rinc)
#frq(db_proc19$us_rinc)
#frq(db_proc19$ve_rinc)
#frq(db_proc19$za_rinc)

# ntile function without NAs
ntile_zero_na <- function(x, n = 10) {
  out   <- rep(NA_real_, length(x))
  is_na <- is.na(x)
  is0   <- !is_na & x == 0
  pos   <- !is_na & x > 0
  out[is0] <- 0
  if (any(pos)) out[pos] <- ntile(x[pos], n)
  out
}

db_proc19 <- db_proc19 %>% 
  mutate_at(vars(5:34), ~ as.numeric(.)) %>% 
 mutate_at(vars(5:34), funs(car::recode(. ,"-9 = NA; -8 = NA; -7 = NA; -2 = NA"))) %>% 
  mutate(
    income_personal = do.call(coalesce, across(5:34)),
    income_decil    = ntile_zero_na(income_personal, 10)
  )

db_proc19 %>% 
  mutate(n_non_na = rowSums(!is.na(pick(5:34)))) %>% 
  count(n_non_na)
frq(db_proc19$income_decil)
```

### Subjective social status
```{r}
frq(db_proc19$ess)
db_proc19$ess <- set_na(db_proc19$ess, na = c(-8,-9), drop.levels = F)
db_proc19$ess <- as.numeric(db_proc19$ess)
frq(db_proc19$ess)
```

### Meritocratic variables
```{r}
frq(db_proc19$merit_effort)

db_proc19 <- db_proc19 %>% 
  mutate(
    across(
      .cols = c(merit_effort),
      .fns = ~ set_na(., na = c(-8,-9), drop.levels = F)
    )
  ) # ok

db_proc19$merit_effort <- sjmisc::rec(db_proc19$merit_effort,rec = "rev")

frq(db_proc19$merit_effort)
```

### Non-meritocratic variables
```{r}
frq(db_proc19$nmerit_wealthy_fam)
frq(db_proc19$nmerit_contacts)
frq(db_proc19$nmerit_educated_parents)
frq(db_proc19$nmerit_political_connection)
frq(db_proc19$nmerit_race)
frq(db_proc19$nmerit_gender)

db_proc19 <- db_proc19 %>% 
  mutate(
    across(
      .cols = c(nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents,
                nmerit_political_connection, nmerit_race, nmerit_gender),
      .fns = ~ set_na(., na = c(-8,-9), drop.levels = F)
    )
  ) # ok

db_proc19$nmerit_wealthy_fam <- sjmisc::rec(db_proc19$nmerit_wealthy_fam,rec = "rev")
db_proc19$nmerit_contacts <- sjmisc::rec(db_proc19$nmerit_contacts,rec = "rev")
db_proc19$nmerit_educated_parents <- sjmisc::rec(db_proc19$nmerit_educated_parents,rec = "rev")
db_proc19$nmerit_political_connection <- sjmisc::rec(db_proc19$nmerit_political_connection,rec = "rev")
db_proc19$nmerit_race <- sjmisc::rec(db_proc19$nmerit_race,rec = "rev")
db_proc19$nmerit_gender <- sjmisc::rec(db_proc19$nmerit_gender,rec = "rev")

frq(db_proc19$nmerit_wealthy_fam)
frq(db_proc19$nmerit_contacts)
frq(db_proc19$nmerit_educated_parents)
frq(db_proc19$nmerit_political_connection)
frq(db_proc19$nmerit_race)
frq(db_proc19$nmerit_gender)

```

### Market justice healthcare
```{r}
frq(db_proc19$just_health)
db_proc19$just_health <- set_na(db_proc19$just_health, na = c(-8, -9), drop.levels = F) 
db_proc19$just_health <- sjmisc::rec(db_proc19$just_health,rec = "rev")
frq(db_proc19$just_health)

```

### Market justice in education
```{r}
frq(db_proc19$just_educ)
db_proc19$just_educ <- set_na(db_proc19$just_educ, na = c(-8,-9), drop.levels = F) 
db_proc19$just_educ <- sjmisc::rec(db_proc19$just_educ,rec = "rev")
frq(db_proc19$just_educ)

```

### Perceived inequality
```{r}
frq(db_proc19$perc_inequality)
db_proc19$perc_inequality <- set_na(db_proc19$perc_inequality, na = c(-8,-9,-4), drop.levels = F)
db_proc19$perc_inequality <- sjmisc::rec(db_proc19$perc_inequality,rec = "rev")
frq(db_proc19$perc_inequality)
```

### Class

```{r}
db_proc19 <- db_proc19 %>%
  mutate(
    isco08 = set_na(isco08, na = c(-9,-8,-4), drop.levels = F),
    emprel = set_na(emprel, na = c(-9,-4), drop.levels = F),
    isco88_4 = isco08to88(isco08),
    selfemp_bin = if_else(emprel %in% c(2,3,4), 1, 0),
    nemploy_rec = case_when(
      emprel == 2         ~ 0,   # self-employed sin empleados
      emprel %in% c(3,4)  ~ 1,   # self-employed con empleados
      TRUE                 ~ NA_real_
    ),
    egp5  = isco88toEGP(isco88_4, n.employees = nemploy_rec,
                         self.employed = selfemp_bin, n.classes = 5)
  )

frq(db_proc19$egp5)
```

```{r}
#| label: class-digclass-2019

db_proc19 <- db_proc19 %>%
  mutate(
    # ISSP 2019 codifica ocupacion con ISCO08 (no ISCO88); DIGCLASS no tiene
    # isco08_to_egp(), por lo que se convierte a ISCO88 con la propia funcion
    # de DIGCLASS (isco08_to_isco88(), distinta de occupar::isco08to88() usada
    # arriba) antes de aplicar el esquema EGP. Para ESeC, DIGCLASS traduce
    # ISCO08 directamente (isco08_to_esec()), sin pasar por ISCO88COM.
    isco08_chr = sprintf("%04d", as.numeric(isco08)),
    isco88_dc_chr = DIGCLASS::isco08_to_isco88(isco08_chr),
    wrksup = set_na(wrksup, na = c(-9, -4), drop.levels = F),
    wrksup_bin = if_else(wrksup == 1, 1, 0),
    # ISSP 2019 no releva el N de empleados de independientes: EMPREL solo
    # distingue "sin empleados" (2) de "con empleados" (3,4). Ver nota sobre
    # el bug de indexacion de DIGCLASS::isco88_to_egp() en el chunk
    # "egp schema" (seccion Cumulative data): para EGP se usa 1 (no 0) en
    # ambos casos de independientes para no caer en la columna equivocada.
    n_employees_egp = if_else(selfemp_bin == 1, 1, 0),
    n_employees_esec = nemploy_rec, # aqui si corresponde 0/1 tal cual (ver chunk "esec schema")
    egp5_dc_code = DIGCLASS::isco88_to_egp(
      isco88_dc_chr,
      self_employed = selfemp_bin,
      n_employees = n_employees_egp,
      n_classes = 5,
      label = FALSE
    ),
    egp5_digclass = factor(
      egp5_dc_code,
      levels = as.character(1:5),
      labels = c("White-collar (I-III)",
                 "Petty bourgeoisie (IV.a-b)",
                 "Farm workers (IV.c+VII.b)",
                 "Skilled workers (V-VI)",
                 "Non-skilled workers (VII.a)")
    ),
    isco08_3 = DIGCLASS::isco08_swap(isco08_chr, from = 4, to = 3),
    esec9_dc = DIGCLASS::isco08_to_esec(
      isco08_3,
      is_supervisor = wrksup_bin,
      self_employed = selfemp_bin,
      n_employees = n_employees_esec
    ),
    # Colapso oficial ESeC-9 -> ESeC-5, ver chunk "esec schema" (seccion
    # Cumulative data) para la referencia (Harrison & Rose 2006, ESeC User
    # Guide, secciones 3.3-3.4).
    esec5_digclass = case_when(
      esec9_dc %in% c("1", "2") ~ "1",
      esec9_dc %in% c("3", "6") ~ "2",
      esec9_dc %in% c("4", "5") ~ "3",
      esec9_dc == "7"           ~ "4",
      esec9_dc %in% c("8", "9") ~ "5"
    ),
    esec5_digclass = factor(
      esec5_digclass,
      levels = as.character(1:5),
      labels = c("Salariat",
                 "Intermediate employees",
                 "Small employers and self-employed",
                 "Lower services, sales and clerical",
                 "Lower technical and routine")
    )
  )

frq(db_proc19$egp5_digclass)
frq(db_proc19$esec5_digclass)
```

## Save

```{r}
#| collapse: false

db_proc19 <- db_proc19 %>%
  select(year, country, iso_country, sex, age, educ, educ_dic, income_decil, ess, just_health, just_educ, perc_inequality, merit_effort, nmerit_wealthy_fam, nmerit_contacts, nmerit_educated_parents, nmerit_political_connection, nmerit_race, nmerit_gender, egp5, egp5_digclass, esec5_digclass, weight)


save(db_proc19, file = here("input/data/proc/db_proc19.RData"))
```

# Final data

## Merge ISSP

```{r}
#| warning: false
#| message: false
dfs_no_labels <- map(list(db_proc87, db_proc92, db_proc99, db_proc09, db_proc19), remove_all_labels)

names(dfs_no_labels) <- c("db_proc87", "db_proc92", "db_proc99", "db_proc09", "db_proc19")
list2env(dfs_no_labels, envir = .GlobalEnv)

db_proc <- bind_rows(
  `1987` = db_proc87,
  `1992` = db_proc92,
  `1999` = db_proc99,
  `2009` = db_proc09,
  `2019` = db_proc19,
  .id = "source_year"
) %>%
  select(-source_year)

db_proc <- tibble::rowid_to_column(db_proc, "id_subject")

head(db_proc)
```

## Save

```{r}
countries_exclude <- c("Argentina", "Belgium", "China", "Estonia", "Ireland",
                        "South Korea", "Suriname", "Thailand", "Turkey", "Ukraine", "Canada")


db_proc <- db_proc %>%
  filter(!country %in% countries_exclude)

save(db_proc, file = here("input/data/proc/issp_proc.RData"))
```

# Cumulative data


## Select

```{r}

db_cumu <- issp_cumu %>%
  select(year, country,
         sex = SEX,
         age = AGE,
         degree = DEGREE,
         income = INCOME,
         ess = TOPBOT,
         merit_effort = v6,
         nmerit_contacts = v7,
         nmerit_race = v9,
         nmerit_gender = v12,
         just_health = v37,
         just_educ = v38,
         perc_inequality = v54,
         isco88 = ISCO88,
         wrksup = WRKSUP,
         selfemp = SELFEMP,
         weight = WEIGHT) %>%
  janitor::clean_names()

```



```{r}
na_codes_cumu <- c(-9,-8,-7,-4,-3,-1)
```

##  Recode and transform

### Year

```{r}
frq(db_cumu$year)
db_cumu$year <- as.numeric(db_cumu$year)
```

### Country

```{r}
db_cumu$country <- as.numeric(db_cumu$country)
db_cumu$country <- countrycode::countrycode(db_cumu$country,
                                             origin = "iso3n",
                                             destination = "country.name")
db_cumu$iso_country <- countrycode::countrycode(db_cumu$country,
                                                 origin = "country.name",
                                                 destination = "iso3c")
frq(db_cumu$country); frq(db_cumu$iso_country)

```

### Sex
```{r}
frq(db_cumu$sex)
db_cumu$sex <- set_na(db_cumu$sex, na = na_codes_cumu, drop.levels = F)
db_cumu <- db_cumu %>%
  mutate(sex = if_else(sex == 1, 'Male', 'Female'),
         sex = factor(sex, levels = c("Male", "Female")))
frq(db_cumu$sex)
```

### Age
```{r}
summary(db_cumu$age)
db_cumu$age <- set_na(db_cumu$age, na = na_codes_cumu, drop.levels = F)
db_cumu$age <- as.numeric(db_cumu$age)
```

### Education
```{r}
frq(db_cumu$degree)
db_cumu$degree <- set_na(db_cumu$degree, na = na_codes_cumu, drop.levels = F)
db_cumu <- db_cumu %>%
  mutate(educ = case_when(degree == 0 ~ "Incomplete primary or less",
                            degree == 1 ~ "Complete primary",
                            degree == 2 ~ "Incomplete secondary",
                            degree == 3 ~ "Complete secondary",
                            degree == 4 ~ "Incomplete university",
                            degree == 5 ~ "Complete university",
                            TRUE ~ NA_character_),
           educ = factor(educ,
                           levels = c("Incomplete primary or less",
                           "Complete primary",
                           "Incomplete secondary",
                           "Complete secondary",
                           "Incomplete university",
                           "Complete university")))

db_cumu <- db_cumu %>%
  mutate(educ_dic = case_when(degree %in% c(0:4) ~ 'Less than universitary',
                              degree == 5 ~ 'Universitary',
                              TRUE ~ NA_character_),
         educ_dic = factor(educ_dic, levels = c("Universitary", "Less than universitary")))
frq(db_cumu$educ)
frq(db_cumu$educ_dic)
```

### Income
```{r}
frq(db_cumu$income)
db_cumu$income <- set_na(db_cumu$income, na = na_codes_cumu, drop.levels = F)
frq(db_cumu$income)
```


### Subjective social status
```{r}
frq(db_cumu$ess)
db_cumu$ess <- set_na(db_cumu$ess, na = na_codes_cumu, drop.levels = F)
db_cumu$ess <- as.numeric(db_cumu$ess)
frq(db_cumu$ess)
```

### Meritocratic variables
```{r}
frq(db_cumu$merit_effort)

db_cumu <- db_cumu %>%
  mutate(
    across(
      .cols = c(merit_effort),
      .fns = ~ set_na(., na = na_codes_cumu, drop.levels = F)
    )
  ) # ok

db_cumu$merit_effort <- sjmisc::rec(db_cumu$merit_effort,rec = "rev")

frq(db_cumu$merit_effort)
```

### Non-meritocratic variables
```{r}
frq(db_cumu$nmerit_contacts)
frq(db_cumu$nmerit_race)
frq(db_cumu$nmerit_gender)

db_cumu <- db_cumu %>%
  mutate(
    across(
      .cols = c(nmerit_contacts, nmerit_race, nmerit_gender),
      .fns = ~ set_na(., na = na_codes_cumu, drop.levels = F)
    )
  ) # ok

db_cumu$nmerit_contacts <- sjmisc::rec(db_cumu$nmerit_contacts,rec = "rev")
db_cumu$nmerit_race <- sjmisc::rec(db_cumu$nmerit_race,rec = "rev")
db_cumu$nmerit_gender <- sjmisc::rec(db_cumu$nmerit_gender,rec = "rev")

frq(db_cumu$nmerit_contacts)
frq(db_cumu$nmerit_race)
frq(db_cumu$nmerit_gender)

```

### Market justice healthcare
```{r}
frq(db_cumu$just_health)
db_cumu$just_health <- set_na(db_cumu$just_health, na = na_codes_cumu, drop.levels = F)
db_cumu$just_health <- sjmisc::rec(db_cumu$just_health,rec = "rev")
frq(db_cumu$just_health)
```

### Market justice in education
```{r}
frq(db_cumu$just_educ)
db_cumu$just_educ <- set_na(db_cumu$just_educ, na = na_codes_cumu, drop.levels = F)
db_cumu$just_educ <- sjmisc::rec(db_cumu$just_educ,rec = "rev")
frq(db_cumu$just_educ)
```

### Perceived inequality
```{r}
frq(db_cumu$perc_inequality)
db_cumu$perc_inequality <- set_na(db_cumu$perc_inequality, na = na_codes_cumu, drop.levels = F)
db_cumu$perc_inequality <- sjmisc::rec(db_cumu$perc_inequality,rec = "rev")
frq(db_cumu$perc_inequality)
```

### Class

```{r}
db_cumu <- db_cumu %>%
  mutate(
    isco88 = set_na(isco88, na = na_codes_cumu, drop.levels = F),
    wrksup = set_na(wrksup, na = na_codes_cumu, drop.levels = F),
    selfemp = set_na(selfemp, na = na_codes_cumu, drop.levels = F),
    selfemp_bin = if_else(selfemp == 1, 1, 0),
    nemploy = NA_real_, # no existe una variable de número de empleados armonizada en el cumulativo
    egp5 = isco88toEGP(isco88, n.employees = nemploy,
                        self.employed = selfemp_bin, n.classes = 5)
  )

frq(db_cumu$egp5)
```

## Save

```{r}
db_cumu <- db_cumu %>%
  select(year, country, iso_country, sex, age, educ, educ_dic, income, ess, just_health, just_educ, perc_inequality, egp5, merit_effort, nmerit_contacts, nmerit_race, nmerit_gender, isco88, wrksup, selfemp_bin, nemploy, weight)

save(db_cumu, file = here("input/data/proc/db_cumu.RData"))

load(here("input/data/proc/db_cumu.RData"))
```

# Criteria for data selection

## Variables available per wave

```{r}
#| label: tbl-vars-cumu
#| tbl-cap: "Missing data by variable and wave (% NA)"
#| tbl-cap-location: top
#| results: asis
#| warning: false

tab_avail_cumu <- db_cumu %>%
  group_by(year) %>%
  summarise(
    n = n(),
    just_health = mean(is.na(just_health)) * 100,
    just_educ = mean(is.na(just_educ)) * 100,
    perc_inequality = mean(is.na(perc_inequality)) * 100,
    merit_effort = mean(is.na(merit_effort)) * 100,
    ess = mean(is.na(ess)) * 100,
    isco88 = mean(is.na(isco88)) * 100,
    wrksup = mean(is.na(wrksup)) * 100,
    .groups = "drop"
  ) %>%
  mutate(across(-c(year, n), ~ round(., 1)),
         year = factor(year))

num_cols_cumu <- setdiff(names(tab_avail_cumu), c("year", "n"))

tab_avail_cumu_fmt <- tab_avail_cumu %>%
  mutate(across(
    all_of(num_cols_cumu),
    ~ ifelse(
      is.na(.x), "",
      ifelse(
        .x == 100,
        kableExtra::cell_spec(sprintf("%.1f", .x), "html", color = "white", background = "darkred"),
        sprintf("%.1f", .x)
      )
    )
  ))

kableExtra::kable(tab_avail_cumu_fmt,
                  format = "html",
                  booktabs = T,
                  escape = F,
                  align = 'c',
                  format.args = list(big.mark = ",")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", full_width = FALSE))

```


## Countries per wave


```{r}
#| label: tbl-country-wave-avail
#| tbl-cap: "Available countries by wave"
#| tbl-cap-location: top
#| results: asis
#| warning: false

tab_country_wave <- db_proc %>%
  distinct(country, year) %>%
  mutate(available = 1L) %>%
  tidyr::pivot_wider(names_from = year, values_from = available, values_fill = 0L) %>%
  arrange(country)

wave_cols <- setdiff(names(tab_country_wave), "country")
tab_country_wave <- tab_country_wave %>%
  select(country, all_of(sort(wave_cols))) %>%
  mutate(Total = rowSums(across(all_of(wave_cols)))) %>%
  janitor::adorn_totals("row")

tab_country_wave_fmt <- tab_country_wave %>%
  mutate(across(
    all_of(wave_cols),
    ~ ifelse(
      .x == 1,
      kableExtra::cell_spec(.x, "html", color = "white", background = "darkred"),
      as.character(.x)
    )
  ))

kableExtra::kable(tab_country_wave_fmt,
                  format = "html",
                  booktabs = T,
                  escape = F,
                  align = 'c') %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", full_width = FALSE))

```

## Suggestions for analysis criteria

- Due to the absence of market justice variables in the 1987 and 1992 waves, we have retained the **last 3 ISSP waves (1999, 2009, 2019)**

- By using the last 3 waves, we have **96 country-waves** in the total sample

- If we apply the selection criterion of including only countries that appear in **at least two** versions of the survey, the number of country-waves decreases to **85** (Argentina, Belgium, Canada, China, Estonia, Ireland, South Korea, Suriname, Thailand, Turkey, and Ukraine are excluded)

```{r}
#| label: clean-countries

countries_exclude <- c("Argentina", "Belgium", "China", "Estonia", "Ireland",
                        "South Korea", "Suriname", "Thailand", "Turkey", "Ukraine", "Canada")

years_exclude <- c("1987", "1992")

db_cumu <- db_cumu %>%
  filter(!country %in% countries_exclude) %>%
  filter(!year %in% years_exclude)
```



# Descriptives

## N by country and wave

```{r}
#| label: tbl-n-country-wave
#| tbl-cap: "N total by country and survey wave"
#| tbl-cap-location: top
#| results: asis
#| echo: false
#| warning: false

tab_n <- db_cumu %>%
  count(country, year, name = "n") %>%
  tidyr::pivot_wider(names_from = year, values_from = n, values_fill = 0) %>%
  select(country, `1999`, `2009`, `2019`) %>%
  mutate(Total = `1999` + `2009` + `2019`) %>%
  arrange(country) %>%
  janitor::adorn_totals("row")

kableExtra::kable(tab_n,
                  format = "html",
                  booktabs = T,
                  align = 'c',
                  format.args = list(big.mark = ",")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", full_width = FALSE))

```

## Class processing (cumulative)

```{r}
#| label: egp schema


db_cumu <- db_cumu %>%
  mutate(
    # ISCO88 4 digitos como string; en este cumulativo el unico codigo con
    # menos de 4 digitos es 110 ("fuerzas armadas"), que pierde el cero inicial
    # al guardarse como numerico (DIGCLASS lo reconoce como "0110")
    isco88_chr = sprintf("%04d", isco88),
    # El cumulativo ISSP no tiene una variable armonizada de N empleados. No se
    # puede usar 0 para todos los casos: DIGCLASS::isco88_to_egp() tiene un bug
    # de indexacion de columnas (case_when de self_employed==1 & n_employees==0
    # apunta a la columna "EGP(0,10+)" en vez de "EGP(1,0)"), por lo que un
    # trabajador independiente sin empleados queda mal clasificado segun su
    # ocupacion como si fuera asalariado de una empresa grande (verificado con
    # los datos de ejemplo del propio paquete: ess$self_employed==1 &
    # ess$emplno==0 nunca cae en la clase "Petty bourgeoisie"). Usar 1 en vez
    # de 0 para independientes esquiva el bug y cae en la columna correcta
    # "EGP(1,0)".
    n_employees_egp = if_else(selfemp_bin == 1, 1, 0),
    egp5_code = DIGCLASS::isco88_to_egp(
      isco88_chr,
      self_employed = selfemp_bin,
      n_employees = n_employees_egp,
      n_classes = 5,
      label = FALSE
    ),
    egp5_digclass = factor(
      egp5_code,
      levels = as.character(1:5),
      labels = c("White-collar (I-III)",
                 "Petty bourgeoisie (IV.a-b)",
                 "Farm workers (IV.c+VII.b)",
                 "Skilled workers (V-VI)",
                 "Non-skilled workers (VII.a)")
    )
  )

frq(db_cumu$egp5_digclass)
```

```{r}
#| label: egp schema - 3 cat

db_cumu <- db_cumu %>%
  mutate(
    # Mismo isco88_chr / n_employees_egp calculados en el chunk "egp schema"
    # (misma logica y mismo workaround del bug de DIGCLASS::isco88_to_egp()
    # descrito ahi); aqui solo se pide el agregado a 3 clases (n_classes = 3),
    # que DIGCLASS obtiene colapsando EGP11 via all_schema$egp11_to_egp3.
    egp3_code = DIGCLASS::isco88_to_egp(
      isco88_chr,
      self_employed = selfemp_bin,
      n_employees = n_employees_egp,
      n_classes = 3,
      label = FALSE
    ),
    egp3_digclass = factor(
      egp3_code,
      levels = as.character(1:3),
      labels = c("Non-manual workers (I-II+IV.a-b)",
                 "Farm workers (IV.c+VII.b)",
                 "Manual workers (V-VI+VII.a)")
    )
  )

frq(db_cumu$egp3_digclass)
```

```{r}
#| label: esec schema


db_cumu <- db_cumu %>%
  mutate(
    wrksup_bin = if_else(wrksup == 1, 1, 0),
    isco88com = DIGCLASS::isco88_to_isco88com(isco88_chr),
    isco88com_3 = DIGCLASS::isco88_swap(isco88com, from = 4, to = 3),
    esec9_digclass = DIGCLASS::isco88com_to_esec(
      isco88com_3,
      is_supervisor = wrksup_bin,
      self_employed = selfemp_bin,
      n_employees = 0, # el cumulativo ISSP no tiene una variable armonizada de N empleados
      full_method = TRUE
    ),
    # Colapso oficial ESeC-9 -> ESeC-5 (Harrison & Rose 2006, ESeC User Guide,
    # secciones 3.3-3.4): DIGCLASS no ofrece un n_classes para ESeC como si
    # hace para EGP, por lo que la agregacion de 9 a 5 clases se aplica aqui
    # siguiendo el modelo de 6 clases oficial y luego combinando clases 5 y 6
    # (tecnicos de baja cualificacion + rutina) en una sola categoria.
    esec5_digclass = case_when(
      esec9_digclass %in% c("1", "2") ~ "1",
      esec9_digclass %in% c("3", "6") ~ "2",
      esec9_digclass %in% c("4", "5") ~ "3",
      esec9_digclass == "7"           ~ "4",
      esec9_digclass %in% c("8", "9") ~ "5"
    ),
    esec5_digclass = factor(
      esec5_digclass,
      levels = as.character(1:5),
      labels = c("Salariat",
                 "Intermediate employees",
                 "Small employers and self-employed",
                 "Lower services, sales and clerical",
                 "Lower technical and routine")
    )
  )

frq(db_cumu$esec5_digclass)
```

```{r}
#| label: tbl-class-valid-cases
#| tbl-cap: "Valid (classifiable) cases by scheme and country"
#| tbl-cap-location: top
#| results: asis
#| echo: false
#| warning: false

tab_class_valid <- db_cumu %>%
  group_by(country) %>%
  summarise(
    n_total = n(),
    n_egp5 = sum(!is.na(egp5_digclass)),
    pct_egp5 = round(n_egp5 / n_total * 100, 1),
    n_esec5 = sum(!is.na(esec5_digclass)),
    pct_esec5 = round(n_esec5 / n_total * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(country)

tab_class_valid <- bind_rows(
  tab_class_valid,
  tab_class_valid %>%
    summarise(country = "Total",
              n_total = sum(n_total),
              n_egp5 = sum(n_egp5),
              pct_egp5 = round(n_egp5 / n_total * 100, 1),
              n_esec5 = sum(n_esec5),
              pct_esec5 = round(n_esec5 / n_total * 100, 1))
)

kableExtra::kable(tab_class_valid,
                  format = "html",
                  booktabs = TRUE,
                  align = 'c',
                  col.names = c("Country", "N total", "N valid", "% valid", "N valid", "% valid")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE) %>%
  kableExtra::add_header_above(c(" " = 2, "EGP-5" = 2, "ESeC-5" = 2)) %>%
  kableExtra::row_spec(nrow(tab_class_valid), bold = TRUE)
```

## Descriptive EGP

### 5 class schema

```{r}
#| label: fig-egp-country
#| fig-cap: "EGP-5 class distribution by country (DIGCLASS)"
#| fig-width: 13
#| fig-height: 20
#| echo: false
#| warning: false

egp_colors_country <- c(
  "White-collar (I-III)"        = "#2a78d6",
  "Petty bourgeoisie (IV.a-b)"  = "#eb6834",
  "Farm workers (IV.c+VII.b)"   = "#1baf7a",
  "Skilled workers (V-VI)"      = "#eda100",
  "Non-skilled workers (VII.a)" = "#e87ba4"
)

tab_egp_country <- db_cumu %>%
  filter(!is.na(egp5_digclass)) %>%
  count(country, egp5_digclass, name = "n") %>%
  group_by(country) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

ggplot(tab_egp_country, aes(x = egp5_digclass, y = pct, fill = egp5_digclass)) +
  geom_col(width = 0.7) +
  facet_wrap(~ country, ncol = 5) +
  scale_fill_manual(values = egp_colors_country, name = "EGP-5 category") +
  scale_y_continuous(limits = c(0, max(tab_egp_country$pct) * 1.1), expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "% within country", title = "EGP-5 class distribution by country") +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "#c3c2b7", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.6, "lines"),
    strip.text = element_text(face = "bold", color = "#0b0b0b", size = 8),
    strip.background = element_blank(),
    plot.title = element_text(color = "#0b0b0b", face = "bold"),
    plot.background = element_rect(fill = "#fcfcfb", color = NA),
    panel.background = element_rect(fill = "#fcfcfb", color = NA),
    legend.position = "bottom"
  )

```

### 3 class schema

```{r}
#| label: fig-egp3-country
#| fig-cap: "EGP-3 class distribution by country (DIGCLASS)"
#| fig-width: 13
#| fig-height: 20
#| echo: false
#| warning: false

egp3_colors_country <- c(
  "Non-manual workers (I-II+IV.a-b)" = "#2a78d6",
  "Farm workers (IV.c+VII.b)"        = "#1baf7a",
  "Manual workers (V-VI+VII.a)"      = "#eda100"
)

tab_egp3_country <- db_cumu %>%
  filter(!is.na(egp3_digclass)) %>%
  count(country, egp3_digclass, name = "n") %>%
  group_by(country) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

ggplot(tab_egp3_country, aes(x = egp3_digclass, y = pct, fill = egp3_digclass)) +
  geom_col(width = 0.7) +
  facet_wrap(~ country, ncol = 5) +
  scale_fill_manual(values = egp3_colors_country, name = "EGP-3 category") +
  scale_y_continuous(limits = c(0, max(tab_egp3_country$pct) * 1.1), expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "% within country", title = "EGP-3 class distribution by country") +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "#c3c2b7", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.6, "lines"),
    strip.text = element_text(face = "bold", color = "#0b0b0b", size = 8),
    strip.background = element_blank(),
    plot.title = element_text(color = "#0b0b0b", face = "bold"),
    plot.background = element_rect(fill = "#fcfcfb", color = NA),
    panel.background = element_rect(fill = "#fcfcfb", color = NA),
    legend.position = "bottom"
  )

```

## Descriptive ESeC

```{r}
#| label: fig-esec-country
#| fig-cap: "ESeC-5 class distribution by country (DIGCLASS)"
#| fig-width: 13
#| fig-height: 20
#| echo: false
#| warning: false

esec_colors_country <- c(
  "Salariat"                            = "#2a78d6",
  "Intermediate employees"              = "#eb6834",
  "Small employers and self-employed"   = "#1baf7a",
  "Lower services, sales and clerical"  = "#eda100",
  "Lower technical and routine"         = "#e87ba4"
)

tab_esec_country <- db_cumu %>%
  filter(!is.na(esec5_digclass)) %>%
  count(country, esec5_digclass, name = "n") %>%
  group_by(country) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

ggplot(tab_esec_country, aes(x = esec5_digclass, y = pct, fill = esec5_digclass)) +
  geom_col(width = 0.7) +
  facet_wrap(~ country, ncol = 5) +
  scale_fill_manual(values = esec_colors_country, name = "ESeC-5 category") +
  scale_y_continuous(limits = c(0, max(tab_esec_country$pct) * 1.1), expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "% within country", title = "ESeC-5 class distribution by country") +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "#c3c2b7", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.6, "lines"),
    strip.text = element_text(face = "bold", color = "#0b0b0b", size = 8),
    strip.background = element_blank(),
    plot.title = element_text(color = "#0b0b0b", face = "bold"),
    plot.background = element_rect(fill = "#fcfcfb", color = NA),
    panel.background = element_rect(fill = "#fcfcfb", color = NA),
    legend.position = "bottom"
  )

```

To select the most appropriate class scheme, we consider two factors: 1) the number of valid cases and 2) the distribution of the classes. 

In the case of the EGP classification scheme, we first tested its 5-class version (see @fig-egp-country), where we observed that White-collar workers are the largest category in all countries, followed by Skilled workers. The Petty Bourgeoisie varies in prevalence depending on the country, while Farm Workers and Unskilled workers are nearly nonexistent in most countries. Due to this overrepresentation of certain classes, we decided to test in additional way the 3-class EGP framework (see @fig-egp3-country). Here, we see a slightly more balanced distribution between non-manual workers and manual workers, while farm workers continue to represent a minimal percentage of the population. 

We then tested the 5-category ESeC framework (see @fig-esec-country). Although the category of salariats is the most prevalent in most countries, the other categories also have levels of frequency that can be useful when considering the analysis phase. Compared to the five-class EGP framework, the ESeC framework is better able to identify the various socioeconomic sectors that exist today.

Therefore, even though the EGP framework has a higher number of valid cases than the ESeC framework (~5%), the latter allows for a better observation of differences in objective status, making it a better alternative for the operationalization of class in international comparisons

## Market justice

```{r}
db_cumu <- db_cumu %>%
  mutate(market_justice = (just_health + just_educ) / 2)
```

```{r}
#| label: tbl-market-justice-country-wave
#| tbl-cap: "Market justice index (mean) by country and survey wave"
#| tbl-cap-location: top
#| results: asis
#| echo: false
#| warning: false

tab_mj <- db_cumu %>%
  group_by(country, year) %>%
  summarise(market_justice = mean(market_justice, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = year, values_from = market_justice) %>%
  select(country, `1999`, `2009`, `2019`) %>%
  mutate_if(is.numeric, ~ round(., 2)) %>%
  arrange(country)

num_cols_mj <- setdiff(names(tab_mj), "country")

tab_mj_fmt <- tab_mj %>%
  mutate(across(
    all_of(num_cols_mj),
    ~ ifelse(
      is.na(.x), "",
      ifelse(
        .x > 2.50,
        kableExtra::cell_spec(sprintf("%.2f", .x), "html", color = "white", background = "darkred"),
        sprintf("%.2f", .x)
      )
    )
  ))

kableExtra::kable(tab_mj_fmt,
                  format = "html",
                  booktabs = T,
                  escape = F,
                  align = 'c') %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", full_width = FALSE))

```


## Social subjective status

```{r}
#| label: fig-ess-wave
#| fig-cap: "Subjective social status (ESS) by ISSP wave"
#| fig-width: 8.5
#| fig-height: 5
#| echo: false
#| warning: false

ess_colors <- c("1999" = "#2a78d6", "2009" = "#eb6834", "2019" = "#1baf7a")

tab_ess <- db_cumu %>%
  filter(!is.na(ess), ess >= 1, ess <= 10) %>%
  mutate(wave = factor(year)) %>%
  count(wave, ess, name = "n") %>%
  group_by(wave) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

label_df <- tab_ess %>% group_by(wave) %>% filter(ess == max(ess)) %>% ungroup()

ggplot(tab_ess, aes(x = ess, y = pct, color = wave, group = wave)) +
  geom_line(linewidth = 0.75) +
  geom_point(size = 2, fill = "#fcfcfb") +
  ggrepel::geom_text_repel(data = label_df, aes(label = wave), hjust = 0, nudge_x = 0.4,
                            direction = "y", segment.color = NA, size = 3.5, fontface = "bold",
                            show.legend = FALSE) +
  scale_color_manual(values = ess_colors, guide = "none") +
  scale_x_continuous(breaks = 1:10, limits = c(1, 11.3), expand = expansion(mult = c(0.02, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
  labs(x = "Subjective social status (1 = bottom, 10 = top)", y = "% within wave",
       title = "Subjective social status (ESS) by ISSP wave") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#e1e0d9"),
    axis.line = element_line(color = "#c3c2b7"),
    axis.text = element_text(color = "#52514e"),
    plot.title = element_text(color = "#0b0b0b", face = "bold"),
    plot.background = element_rect(fill = "#fcfcfb", color = NA),
    panel.background = element_rect(fill = "#fcfcfb", color = NA)
  )
```

### Meritocratic items

#### Effort

```{r}
#| label: fig-merit-effort-wave
#| fig-cap: "Mean merit_effort by ISSP wave"
#| fig-width: 6.5
#| fig-height: 5.3
#| echo: false
#| warning: false

tab_me_1999 <- db_proc %>%
  # merit_effort (cumulative v6, "how important is hard work") is 100% NA for
  # 1999 in db_cumu: ese item no se administro en el modulo de 1999. Se usa en
  # su lugar el item mas cercano disponible ese ano ("people get rewarded for
  # their effort", v6 del archivo original 1999), ya calculado en db_proc.
  filter(year == 1999, !is.na(merit_effort)) %>%
  summarise(wave = "1999", mean_effort = mean(merit_effort))

tab_me_cumu <- db_cumu %>%
  filter(!is.na(merit_effort)) %>%
  mutate(wave = as.character(year)) %>%
  group_by(wave) %>%
  summarise(mean_effort = mean(merit_effort), .groups = "drop")

tab_me <- bind_rows(tab_me_1999, tab_me_cumu) %>%
  mutate(wave = factor(wave, levels = c("1999", "2009", "2019")))

ggplot(tab_me, aes(x = wave, y = mean_effort)) +
  geom_col(fill = "#2a78d6", width = 0.55) +
  geom_text(aes(label = sprintf("%.2f", mean_effort)), vjust = -0.6, size = 4, color = "#0b0b0b", fontface = "bold") +
  scale_y_continuous(limits = c(0, max(tab_me$mean_effort) * 1.18), expand = expansion(mult = c(0, 0))) +
  labs(x = NULL, y = "Mean merit_effort (1-5)",
       title = "Belief in the importance/reward of effort, by ISSP wave",
       caption = "Note: 1999 measures a different item (\"people get rewarded for their effort\", agree-disagree)\nthan 2009/2019 (\"how important is hard work for getting ahead\", ISSP cumulative v6). Not directly comparable.") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "#e1e0d9"),
    axis.line.x = element_line(color = "#c3c2b7"),
    axis.text = element_text(color = "#52514e"),
    plot.title = element_text(color = "#0b0b0b", face = "bold"),
    plot.caption = element_text(color = "#898781", size = 8, hjust = 0, face = "italic"),
    plot.background = element_rect(fill = "#fcfcfb", color = NA),
    panel.background = element_rect(fill = "#fcfcfb", color = NA)
  )

```
